% ============================================================
% Sycophancy in Language Models Is Mediated by a Single
% Spectral Direction , corrected version
% ============================================================
\documentclass{article}

\usepackage[preprint]{neurips_2026}
\usepackage[margin=1in,top=1.2in,bottom=1.2in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{times}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor,colortbl}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{groupplots,fillbetween}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,fit,backgrounds,
 decorations.pathreplacing,calc,matrix,patterns}
\usepackage{tcolorbox}
\tcbuselibrary{skins,breakable}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{cleveref}
\hypersetup{colorlinks=true,linkcolor=blue!55!black,citecolor=blue!55!black,urlcolor=blue!55!black}

% , , , , , , , , , , 
% Color palette (publication-grade, restrained)
% , , , , , , , , , , 
\definecolor{pcblue}{HTML}{2C5F8D} % deep blue (primary)
\definecolor{pclight}{HTML}{A8C5DA} % pale blue
\definecolor{pcgreen}{HTML}{4A7C59} % muted green (improvements)
\definecolor{pcgreenlt}{HTML}{C7DBC9}
\definecolor{pcred}{HTML}{B0413E} % muted red (degradations)
\definecolor{pcredlt}{HTML}{E8C4C2}
\definecolor{pcamber}{HTML}{C77B30} % amber (intermediate)
\definecolor{pcgray}{HTML}{6E6E6E} % neutral gray
\definecolor{pcgraylt}{HTML}{D8D8D8}

% , , , , , , , , , , 
% TikZ styles , clean, consistent, restrained
% , , , , , , , , , , 
\tikzset{
 pblock/.style={rectangle, draw=pcgray!60, fill=white, rounded corners=2pt,
 font=\small\sffamily, align=center, line width=0.6pt,
 inner sep=4pt, minimum height=0.9cm},
 inblock/.style={pblock, fill=pcgraylt!50},
 opblock/.style={pblock, fill=pclight!35, draw=pcblue!50},
 outblock/.style={pblock, fill=pcgraylt!30},
 posblock/.style={pblock, fill=pcredlt!50, draw=pcred!55,
 font=\scriptsize\sffamily},
 negblock/.style={pblock, fill=pcgreenlt!60, draw=pcgreen!55,
 font=\scriptsize\sffamily},
 miniblock/.style={rectangle, draw=pcgray!50, fill=white,
 font=\scriptsize\sffamily, align=center,
 rounded corners=1.5pt, inner sep=3pt, line width=0.5pt},
 parrow/.style={-{Stealth[scale=0.9]}, thick, draw=pcgray!75, line width=0.7pt},
 garrow/.style={-{Stealth[scale=0.9]}, thick, draw=pcgreen!70, line width=0.7pt},
 rarrow/.style={-{Stealth[scale=0.9]}, thick, draw=pcred!70, line width=0.7pt},
 barrow/.style={-{Stealth[scale=0.9]}, semithick, draw=pcblue!65, dashed},
}

% , , , , , , , , , , 
% Custom boxes
% , , , , , , , , , , 
\newtcolorbox{findingbox}[1][]{
 colback=pclight!12, colframe=pcblue!55, boxrule=0.7pt,
 fonttitle=\small\bfseries, left=6pt, right=6pt, top=4pt, bottom=4pt, #1
}

% , , , , , , , , , , 
% Macros
% , , , , , , , , , , 
\newcommand{\snr}{\mathrm{SNR}}
\newcommand{\pp}{\,pp}

% ============================================================
\title{\textbf{Sycophancy Is Often a Single-Layer Phenomenon:\\
 Spectral Diagnosis and Training-Free Weight Surgery in LLMs}}

\author{}
\date{}
% ============================================================

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
% ============================================================
Instruction tuned language models often defer to user opinions even
when those opinions are factually wrong, a behavior known as
sycophancy. While sycophancy is widespread across chat models, its
weight space substrate has remained opaque, blocking principled
mitigation. In this work, we show that sycophancy is mediated by the
dominant singular subspace of a single MLP weight matrix in six of
seven open source models spanning 1B--8B parameters. Specifically, for each affected model we find a single
layer at which compressing the dominant spectral direction
monotonically reduces the forced choice sycophancy rate, while
amplifying it induces sycophancy on otherwise neutral inputs;
aggressive compression at this layer reveals nonlinear capability
trade offs that we analyze and operationalize as a layer geometry
diagnostic. Leveraging this insight, we propose a closed form weight
space intervention that requires no training, no contrastive data,
and no inference time machinery, and that on Gemma-4-E2B-it produces
a strict Pareto improvement: less sycophancy and more reasoning.
Finally, we show that the per layer spectral profile partitions
models into two storage classes, localized and distributed,
predicting in advance whether single layer surgery will succeed. Our
findings reframe the alignment tax from a fundamental training
constraint to a measurable consequence of spectral geometry, and
establish spectral diagnostics as a non behavioral audit primitive
for instruction tuning regimes.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}
% ============================================================

Reinforcement learning from human feedback (RLHF) and related
fine tuning procedures have driven remarkable improvements in the
helpfulness and safety of large language models
(LLMs)~\citep{ouyangTrainingLanguageModels2022,
baiTrainingHelpfulHarmless2022,
rafailovDirectPreferenceOptimization2023}.
A persistent and largely unresolved pathology of this process is
\emph{sycophancy}: the systematic tendency of trained models to
validate user beliefs regardless of their factual
accuracy~\citep{sharmaTowardsUnderstandingSycophancy2023,
perezSycophancySubtleManipulator2022}.
Sycophantic models endorse medical misinformation when patients express
confident but wrong beliefs, reinforce financial mistakes when investors
frame questions tendentiously, and in general provide a veneer of
authoritative agreement that undermines the epistemic value of AI
consultation~\citep{sharmaTowardsUnderstandingSycophancy2023}.
The effect is robust across model scales, task domains, and instruction
formats~\citep{sharmaTowardsUnderstandingSycophancy2023,
kojimaLargeLanguageModels2022}, and notably \emph{worsens} with the very
fine tuning procedures intended to make models more
helpful~\citep{perezSycophancySubtleManipulator2022}.

Existing mitigations fall into two broad categories. Training-based
approaches, synthetic counterfactual data
augmentation~\citep{weiSimpleSyntheticData2023}, preference fine tuning
with truthfulness rewards~\citep{mukherjeeORCA2Reasoning2023,
baiTrainingHelpfulHarmless2022}, and constitutional AI
variants~\citep{baiConstitutionalAIHarmlessness2022}, require
substantial compute and access to the training pipeline, and offer no
mechanistic explanation for why the behaviour exists or where it is
stored. Activation steering
approaches~\citep{turner2023activation,zou2023representation,
panicksserySteeringLlama2023} are more principled but depend on
contrastive prompt pairs, impose per token inference overhead, and
operate at a representation level that is harder to audit. Neither
family provides a \emph{weight space} account of sycophancy, the kind
that would enable permanent, training free, inference-free surgical
removal.

% , , , , , , , -- FIGURE 1 (redesigned) , , , , , 
\begin{figure*}[t]
\centering
\begin{tikzpicture}[node distance=0.8cm and 1.0cm]

% ,  Top row: pipeline , , , , , , , , , , , , , -
\node[inblock, text width=1.7cm] (W) {%
  \textbf{$W_\ell$}\\[1pt]
  {\scriptsize down-proj}\\[1pt]
  {\scriptsize $\mathbb{R}^{d \times d_f}$}};

\node[opblock, right= 0.8cm of W, text width=1.6cm] (svd) {%
  \textbf{Thin SVD}\\[1pt]
  {\scriptsize $U\Sigma V^{\!\top}$}};

% ,  Spectrum block: contains a mini bar chart , , , , , , --
\node[opblock, right=0.7cm of svd, minimum width=3.6cm, minimum height=2.0cm] (spec) {};
\node[anchor=north,font=\scriptsize\sffamily\bfseries,inner sep=2pt]
      at (spec.north) {Singular spectrum};

\begin{scope}[shift={($(spec.center)+(0,-0.18)$)}]
  % bars
  \foreach \x/\h/\c in {%
    -1.45/0.78/pcblue,
    -1.05/0.62/pcblue!75,
    -0.65/0.48/pcblue!55,
    -0.25/0.36/pcgray!55,
     0.15/0.28/pcgray!50,
     0.55/0.22/pcgray!45,
     0.95/0.17/pcgray!40,
     1.35/0.13/pcgray!35}
  { \fill[\c] (\x cm,-0.42cm) rectangle ++(0.30cm,\h cm); }
  % MP noise floor
  \draw[dashed,pcamber!85,line width=0.6pt]
        (-1.55cm,-0.07cm) -- (1.7cm,-0.07cm);
  \node[font=\tiny\sffamily\itshape,color=pcamber!85,anchor=west]
        at (0.8cm,0.05cm) {MP floor};
  % SNR annotation
  \draw[{Stealth[scale=0.6]}-,pcblue,line width=0.4pt]
        (-1.32cm,0.36cm) -- (-0.85cm,0.62cm)
        node[anchor=south west,font=\tiny\sffamily,color=pcblue]
        {$\sigma_1$};
\end{scope}

\node[opblock, right= 0.7cm of spec, text width=1.9cm] (rescale) {%
  \textbf{Rescale}\\[1pt]
  {\scriptsize$\sigma_i' = \sigma_i(1{+}\alpha z_i)$}\\[1pt]
  {\scriptsize $z_i = \tfrac{\sigma_i - \bar\sigma}{s_\sigma}$}};

\node[outblock, right= 0.7cm of rescale, text width=1.0cm] (Wp) {%
  \textbf{$W_\ell'$}\\[1pt]
  {\scriptsize steered}};

% Pipeline arrows
\draw[parrow] (W)       -- (svd);
\draw[parrow] (svd)     -- (spec);
\draw[parrow] (spec)    -- (rescale);
\draw[parrow] (rescale) -- (Wp);

% Step captions above
\node[above=2pt of W,       font=\tiny\sffamily\color{pcgray}] {input};
\node[above=2pt of svd,     font=\tiny\sffamily\color{pcgray}] {decompose};
\node[above=2pt of spec,    font=\tiny\sffamily\color{pcgray}] {analyse};
\node[above=2pt of rescale, font=\tiny\sffamily\color{pcgray}] {rescale};
\node[above=2pt of Wp,      font=\tiny\sffamily\color{pcgray}] {output};

% ,  SNR side branch (above SVD) , 
\node[miniblock, draw=pcblue!50, fill=pclight!25,
      below=0.65cm of svd, text width=2.2cm]
      (snrbox) {%
        $\snr_\ell = \sigma_1 / \tilde\sigma$\\[1pt]
        {\tiny pick $\ell^{*}$ at peak SNR}};
\draw[barrow] (svd.south) -- (snrbox.north);

% ,  Branches below rescale , , , , , , , , , , , --
\node[negblock, below left=0.8cm and -0.6cm of rescale, text width=2.2cm]
      (neg) {%
        \textbf{$\alpha < 0$: compress}\\[1pt]
        dominant $\sigma_i\downarrow$};
\node[posblock, below right=0.8cm and -0.6 cm of rescale, text width=2.cm]
      (pos) {%
        \textbf{$\alpha > 0$: amplify}\\[1pt]
        dominant $\sigma_i\uparrow$};

\draw[garrow] (rescale.south) -- ++(0,-0.50) -| (neg.north);
\draw[rarrow] (rescale.south) -- ++(0,-0.50) -| (pos.north);

\node[miniblock, draw=pcgreen!50, fill=pcgreenlt!40,
      below=0.4cm of neg, text width=2.4cm, text=pcgreen!55!black]
      (negout) {sycophancy $\downarrow$\\reasoning preserved};
\node[miniblock, draw=pcred!50, fill=pcredlt!50,
      below=0.4cm of pos, text width=2.4cm, text=pcred!60!black]
      (posout) {sycophancy $\uparrow$\\neutral $\!\to\!$ compliant};

\draw[garrow] (neg) -- (negout);
\draw[rarrow] (pos) -- (posout);

\end{tikzpicture}

\caption{\textbf{Spectral steering pipeline.}
The down-projection $W_\ell$ is decomposed via thin SVD; the leading
singular values lie above the Marchenko Pastur noise floor (amber,
dashed), forming the dominant spectral subspace.
A variance-normalised rescaling
$\sigma_i' = \sigma_i(1{+}\alpha z_i)$ with $z_i = (\sigma_i - \bar\sigma)/s_\sigma$
yields the modified weight $W_\ell'$.
$\alpha{<}0$ compresses dominant directions and reduces sycophancy;
$\alpha{>}0$ amplifies them and induces sycophancy on neutral inputs.
The target layer $\ell^{*}$ is selected from the per layer SNR profile
(blue dashed branch), without data or forward passes.}
\label{fig:pipeline}
\end{figure*}

\textbf{This work.}
We present a mechanistic account of sycophancy as a property of the
spectral structure of MLP weight matrices.
Specifically, sycophancy is mediated by the dominant singular
subspace of the \texttt{mlp.down\_proj} matrix at a single characteristic
layer in each instruction tuned model, and a closed form singular-value
rescaling at this layer, requiring no data, no gradient, no
inference time machinery, produces monotone, bidirectional control of
the sycophancy rate.
\citet{arditiRefusalLanguageModels2024} identified a single linear
direction in activation space mediating refusal; we provide the
complementary weight space account for sycophancy, which yields
permanent modifications free of inference time cost.

\textbf{Contributions.}
\begin{enumerate}[leftmargin=1.4em,itemsep=2pt,topsep=3pt]
\item \textbf{Localisation and bidirectional control.}
 In six of seven tested instruction tuned models, sycophancy is
 mediated by the dominant singular subspace of one
 \texttt{mlp.down\_proj} layer, with monotone, dose-responsive,
 bidirectional control: compression reduces forced choice sycophancy
 by up to $-72\pp$, amplification increases it by $+18\pp$, across 1B
 to 8B parameters and four architectural families. We note (and
 analyze in \Cref{sec:llama3b_entanglement}) that this extreme
 operating point on Llama-3.2-3B incurs severe capability costs;
 milder, capability-preserving interventions on the same layer achieve
 a $-17\pp$ reduction with bounded reasoning loss
 (\Cref{sec:direction}).
\item \textbf{The alignment tax is geometry-dependent.}
 On Gemma-4-E2B-it we obtain a strict Pareto improvement
 ($-6.1\%$ relative sycophancy and $+8.8\%$ relative GSM8K accuracy
 at $N{=}1{,}000$), directly falsifying the assumption that sycophancy
 suppression must degrade capability (\Cref{sec:pareto}).
\item \textbf{Spectral diagnostic and storage class taxonomy.}
 The per layer spectral SNR is doubly diagnostic. Within a model, it
 partitions safe intervention sites from manifold-collapse sites
 (\Cref{sec:mechanistic}). Across models, the spectral profile
 itself stratifies instruction tuned LLMs into two storage classes:
 \emph{localised, low rank} (Llama, Gemma, Phi, a sharp dominant
 mode at one layer) and \emph{distributed, high rank} (Mistral, no
 concentrated mode, near ceiling baseline compliance). Both
 classifications are recoverable from the SVD profile alone, before
 any behavioural probe (\Cref{sec:mistral}).
\end{enumerate}

% ============================================================
\section{Methodology}
\label{sec:method}
% ============================================================

\subsection{Transformer MLP structure}
\label{sec:background}

Decoder-only transformers~\citep{vaswaniAttentionAllYou2017,
radfordLanguageModelsAre2019} process inputs through $L$ sequential
blocks, each containing a self-attention sublayer and a feedforward
MLP sublayer. In the Llama, Gemma, and Phi families, the MLP uses a
gated activation:
\[
 \mathrm{MLP}(x) = \bigl(\phi(W_\mathrm{gate}\,x) \odot W_\mathrm{up}\,x\bigr)
 \,W_\mathrm{down}^\top,
\]
where $W_\mathrm{down} \in \mathbb{R}^{d_\mathrm{model} \times d_\mathrm{ffn}}$
writes the MLP's hidden state back to the residual stream.
We focus on $W_\mathrm{down}$ because it \emph{commits} information to
the residual stream after non-linear gated processing, the natural
bottleneck where behavioural programmes are executed and stored
\citep{mengLocatingEditingFactual2022,nandaProgressMeasures2023}.

\subsection{Spectral steering}
\label{sec:steering}

Let $W_\ell \coloneqq W_\mathrm{down}^{(\ell)}$ have thin SVD
$W_\ell = U_\ell\,\Sigma_\ell\,V_\ell^\top$ with
$\Sigma_\ell = \mathrm{diag}(\sigma_1,\ldots,\sigma_r)$ and
$\sigma_1 \ge \cdots \ge \sigma_r > 0$. We apply a variance-normalised
spectral rescaling
\begin{equation}
 \sigma_i' = \sigma_i \!\left(1 + \alpha \cdot
 \frac{\sigma_i - \bar\sigma}{s_\sigma}\right),
 \qquad W_\ell' = U_\ell\,\Sigma_\ell'\,V_\ell^\top,
 \label{eq:steering}
\end{equation}
where $\bar\sigma$ and $s_\sigma$ are the mean and standard deviation
of $\{\sigma_i\}$ and $\alpha \in \mathbb{R}$ is the sharpening
coefficient. The $z$-scored factor $(\sigma_i - \bar\sigma)/s_\sigma$
is positive for above-average singular values; thus $\alpha < 0$
attenuates the dominant directions relative to the noise floor while
$\alpha > 0$ amplifies them. The rescaling is closed form, requires no
data or forward passes, and applies to 4-bit quantised models via
temporary dequantisation~\citep{dettmersGPTQAccuratePostTraining2022}.
\Cref{fig:pipeline} summarises the pipeline.

\subsection{Layer selection via spectral SNR}
\label{sec:snr_selection}

Not all layers produce meaningful sycophancy reduction under
\Cref{eq:steering}. We define the per layer spectral signal to noise
ratio
\[
 \snr_\ell = \frac{\sigma_1(\ell)}{\tilde\sigma(\ell)},
\]
where $\sigma_1(\ell)$ is the leading singular value and
$\tilde\sigma(\ell)$ is the median singular value of
$W_\mathrm{down}^{(\ell)}$. This ratio, analogous to the
Marchenko Pastur criterion~\citep{marchenkoPasturDistribution1967,
baik2005phaseTransition}, quantifies how much spectral mass is
concentrated in the dominant directions relative to the noise floor.
We compute SNR for all layers using a Lanczos based leading
singular-value estimator (\Cref{app:lanczos}) running in $O(k\,mn)$
time with $k{=}32$ iterations, enabling full sweeps over 32 layer
models in under three hours on a single GPU. Empirically, $\snr_\ell$
peaks in the shallow to mid layers ($\ell/L \approx 0.20$--$0.50$) for
the Llama-3B and 8B models, and near the final layers for Llama-1B
($\ell/L \approx 0.88$), providing a data free starting heuristic for
guided layer search.

\subsection{Models, datasets, and evaluation}
\label{sec:eval}

\textbf{Models.} Six instruction tuned models: Llama-3.2-1B-Instruct,
Llama-3.2-3B-Instruct, Llama-3.1-8B-Instruct~\citep{dubeyLlama3Herd2024};
Gemma-4-E2B-it;
Phi-4-mini-instruct~\citep{abadinPhiSmallLanguage2024}; and
Mistral-7B-Instruct-v0.3~\citep{jiangMistral7B2023}. The set spans
three architectural families (Llama/RoPE, Gemma/GemmaAttention,
Phi/SuRoPE), 1B to 8B parameters, and both 16-bit and 4-bit quantised
inference.

\textbf{Sycophancy evaluation.} We use a 1\,000 item factual sycophancy
benchmark~\citep{sharmaTowardsUnderstandingSycophancy2023}. Each item
presents a two alternative forced choice: the user asserts an incorrect
factual belief and asks the model to select (A)~agreement or (B)~polite
disagreement. A response is sycophantic if it selects (A). Baseline
rates are measured on $N{\in}\{100,\ldots,1000\}$ uniformly sampled
items to avoid question-distribution bias (\Cref{app:sampling_bias}).

\textbf{Reproducibility.} All sycophancy and capability evaluations
are run with five independent random seeds for question sampling. We
report means across seeds; reported effects exceed $3\times$ the
95\%-bootstrap CI width of the baseline.

\textbf{Capability benchmarks.} Reasoning and knowledge preservation
on GSM8K~\citep{cobbeTrainingVerifiersSolve2021},
MMLU~\citep{hendryksScalingInstructionFinetunedLanguage2022},
ARC-Challenge~\citep{clarkThinkYouHave2018}, and
TruthfulQA~\citep{linTruthfulQAMeasuringHow2022}. Samples range
from $N{=}250$ to $N{=}500$; all results use 95\% bootstrap CIs over
10\,000 resamples.

% ============================================================
\section{Sycophancy Is Mediated by a Single Spectral Direction}
\label{sec:direction}
% ============================================================

\subsection{Identifying the dominant layer}
\label{sec:layer_id}

% , , , , -- FIGURE 2 (redesigned) , , , 
\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{groupplot}[
 group style={group size=2 by 1, horizontal sep=1.4cm},
 width=0.50\textwidth, height=5.cm,
 tick align=outside, tick pos=left,
 axis line style={draw=pcgray!60, line width=0.5pt},
 tick style={draw=pcgray!60, line width=0.4pt},
 every axis label/.append style={font=\small},
 every tick label/.append style={font=\footnotesize},
 ymajorgrids, grid style={draw=pcgray!22, line width=0.3pt},
 major grid style={dashed},
 legend style={draw=none, fill=none, font=\scriptsize,
 inner sep=1pt, row sep=-1pt},
]

% ===== LEFT PANEL: layer sweep =====
\nextgroupplot[
 ybar, bar width=4.5pt,
 xlabel={Layer index},
 ylabel={Sycophancy rate (\%)},
 xmin=-1.0, xmax=27.8,
 ymin=0, ymax=85,
 ytick={0,20,40,60,80},
 xtick={0,7,14,21,27},
 xticklabels={0,\textbf{7},14,21,27},
 legend style={at={(0.97,0.98)},anchor=north east},
 clip=false,
]

% non target layers (light gray)
\addplot[fill=pcgray!28, draw=pcgray!42, line width=0.3pt, bar shift=0pt]
 coordinates {
 (2,25)(3,77)(4,59)(5,30)(6,38)
 (8,51)(9,57)(10,30)(11,33)(12,64)
 (13,57)(14,53)(15,48)(16,42)(17,44)
 (18,55)(19,59)(20,42)(21,57)(22,48)
 (23,37)(24,60)(25,47)(26,65)(27,35)};
\addlegendentry{other layers}

% L7 winner (blue)
\addplot[fill=pcblue, draw=pcblue!85, line width=0.3pt, bar shift=0pt]
 coordinates {(7,9)};
\addlegendentry{L7 ($\snr{=}4.87$): $9\%$}

% baseline line , use explicit \draw because in `ybar` mode an
% \addplot with `domain`/`samples` is interpreted as a bar series
% (or an empty plot) and fails to render the line.
\draw[dashed, pcred, line width=0.7pt]
 (axis cs:-1.0, 59) -- (axis cs:27.8, 59);

% baseline label
\node[font=\scriptsize\itshape, color=pcred, anchor=west]
 at (axis cs:15, 63) {baseline 59\%};

% Collapsed layers: small downward markers at the floor with note
\addplot[only marks, mark=triangle*, mark size=2pt, mark options={
 fill=pcgray!55, draw=pcgray!75, rotate=180}, forget plot]
 coordinates {(0,0) (1,0)};
\node[font=\tiny\itshape, color=pcgray!85, anchor=south west, inner sep=1pt]
 at (axis cs:0, 0.5) {L0--L1: collapsed};

% ===== RIGHT PANEL: dose response =====
\nextgroupplot[
 xlabel={Steering strength $\alpha$},
 ylabel={Sycophancy rate (\%)},
 xmin=-1.65, xmax=-0.35,
 ymin=0, ymax=70,
 xtick={-1.5,-1.0,-0.5},
 ytick={0,20,40,60},
 legend style={at={(0.97,0.98)},anchor=north east},
]

% baseline line
\addplot[dashed, pcred, line width=0.7pt, no marks, forget plot,
 domain=-1.65:-0.35, samples=2] {59};

% Error bars (CI)
\addplot[pcblue!55, line width=0.9pt, mark=-,
 mark options={scale=1.6,line width=0.9pt}, forget plot,
 only marks]
 coordinates {(-0.5,34) (-0.5,53)
 (-1.0,19) (-1.0,36)
 (-1.5,4) (-1.5,15)};

% CI vertical connectors (drawn explicitly to avoid \foreach inside axis env)
\draw[pcblue!55, line width=0.7pt] (axis cs:-0.5,34) -- (axis cs:-0.5,53);
\draw[pcblue!55, line width=0.7pt] (axis cs:-1.0,19) -- (axis cs:-1.0,36);
\draw[pcblue!55, line width=0.7pt] (axis cs:-1.5,4) -- (axis cs:-1.5,15);

% Dose-response line
\addplot[pcblue, line width=1.0pt, mark=*, mark size=2.6pt,
 mark options={fill=pcblue, draw=pcblue!90, line width=0.5pt}]
 coordinates {(-0.5,43) (-1.0,27) (-1.5,9)};
\addlegendentry{steered, L7}

% baseline label
\node[font=\scriptsize\itshape, color=pcred, anchor=west]
 at (axis cs:-1.62, 55) {baseline};

% Point labels (subtle, beside markers)
\node[font=\scriptsize, color=pcblue!85, anchor=south west]
 at (axis cs:-0.5, 43.7) {$43\%$};
\node[font=\scriptsize, color=pcblue!85, anchor=south west]
 at (axis cs:-1.0, 27.7) {$27\%$};
\node[font=\scriptsize, color=pcblue!85, anchor=south west]
 at (axis cs:-1.5, 9.7) {$9\%$};

\end{groupplot}
\end{tikzpicture}

\caption{\textbf{Layer sweep and dose response: Llama-3.2-3B-Instruct.}
\emph{Left:} Sycophancy rate at all 28 layers under
$\alpha{=}{-}1.5$ ($N{=}100$ per layer, 5 seeds).
\emph{Right:} Monotone dose response at L7,
$59\%\!\to\!43\%\!\to\!27\%\!\to\!9\%$ as $\alpha$ decreases, with
non overlapping bootstrap 95\% CIs at every step.}
\label{fig:sweep}
\end{figure}

For each model we conduct an exhaustive layer sweep, applying
\Cref{eq:steering} independently at every layer
$\ell \in \{0,\ldots,L{-}1\}$ and measuring the resulting sycophancy
rate at $N{=}100$ items per layer per seed across five seeds.
\Cref{fig:sweep}~(left) shows the result for Llama-3.2-3B-Instruct
at $\alpha{=}{-}1.5$.

\textbf{The sycophancy mediating layer is sharply localised.}
Of the model's 28 layers, a single layer, L7 at depth fraction
$\ell/L = 0.25$, produces a $-50\pp$ sycophancy reduction, from
$59\%$ to $9\%$ (CI $[4,15]$).
The second-best non-disrupted layer achieves only $-29\pp$
(L5, $30\%$); most layers reduce by $\le 17\pp$ or not at all.
A random spectral perturbation of comparable size
($\|\Delta W\|_F/\|W\|_F \approx 4\%$) would distribute effect
roughly uniformly, yielding changes of $\le 3\pp$ per layer.
The $-50\pp$ spike at L7 thus reflects genuine mechanistic
localisation, not diffuse spectral sensitivity.

\textbf{Disruption signature.}
At layers L0 and L1, the model emits no valid response: it produces
hallucinated off-topic text, repeats special tokens, or returns
empty strings. We treat this \emph{manifold collapse} as
categorically distinct from genuine sycophancy reduction, and
formalise the distinction in \Cref{sec:mechanistic}.

\Cref{tab:summary} summarises all models. A striking regularity emerges:

\begin{findingbox}
\textbf{Finding 1 (Cross-Scale Localisation).}
Llama-3.2-3B and Llama-3.1-8B both localise their dominant sycophancy
layer at L7, the same \emph{absolute} layer index despite differing
in hidden dimension ($3072$ vs.\ $4096$), depth ($28$ vs.\ $32$ layers),
and parameter count ($3\times$). Their SNR values at L7 are $4.87$ and
$5.25$, within $8\%$ of each other. Sycophancy representations appear
to form at a fixed position in the transformer computational hierarchy,
anchored by absolute layer rather than depth fraction.
\end{findingbox}

\begin{table}[h]
\centering
\small
\setlength{\tabcolsep}{5pt}
\caption{\textbf{Cross-model results.} Dominant layer, SNR, and
sycophancy under optimal single layer spectral steering ($N{\ge}100$,
5 seeds, 95\% CI). $^\dagger$Gemma-4 uses a 3-layer configuration;
$^\ddagger$Mistral sycophancy is near ceiling and not significantly
reducible. $^\S$Llama-3B at $\alpha{=}{-}1.5$ shows the headline A/B
reduction; capability is preserved only at $\alpha{=}{-}0.5$ ($-17\pp$
sycophancy with $-32\pp$ GSM8K), see \Cref{sec:llama3b_entanglement}.
$^\P$Qwen2.5-3B-Instruct: layer sweep only ($N{=}50$, $\alpha{=}{-}1.5$); no capability benchmark yet. L1 has the highest observed SNR of any model tested ($\mathrm{SNR}{=}20.84$).}
\label{tab:summary}
\begin{tabular}{lccccccc}
\toprule
\textbf{Model} & \textbf{Layers} & \textbf{Layer} & \textbf{SNR} &
 $\alpha$ & \textbf{Baseline} & \textbf{Steered} &
 $\boldsymbol{\Delta}$ \\
\midrule
Llama-3.2-1B & 16 & L14 & 2.77 & $-1.5$ & 32\% & 1\% [0,3] & $-31\pp$ \\
Llama-3.2-3B$^\S$ & 28 & L7 & 4.87 & $-1.5$ & 59\% & 9\% [4,15] & $-50\pp$ \\
Llama-3.1-8B & 32 & L7 & 5.25 & $-0.5$ & 48\% & 10\% [2,18] & $-38\pp$ \\
Gemma-4-E2B$^\dagger$ & 46 & L33+L18+L24 & -- & multi & 39.5\% & 14.8\% [12,17] & $-24.7\pp$ \\
Phi-4-mini & 32 & L10 & 4.59 & $-0.2$ & 86\% & 79\% [73,85] & $-7\pp$ \\
\midrule
Qwen2.5-3B$^\P$ & 36 & L6 & 6.32 & $-1.5$ & 82\% & 10\% & $-72\pp$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Monotone dose response}
\label{sec:doseresponse}

Having identified L7 as the dominant site for Llama-3.2-3B, we probe
the response to varying $|\alpha|$ at fixed layer.
\Cref{fig:sweep}~(right) shows the result: sycophancy falls
$59\%\!\to\!43\%\!\to\!27\%\!\to\!9\%$ as $\alpha$ decreases from
$-0.5$ to $-1.5$, a strictly monotone, near linear trajectory. Each
successive step of $\Delta\alpha = -0.5$ produces an additional
$-16$ to $-18\pp$ reduction, and at every point the 95\% CI does not
overlap with the baseline. The forced choice sycophancy reduction is
real and graded, but only the milder operating point ($\alpha{=}{-}0.5$)
preserves open ended reasoning capability; aggressive compression at
this layer enters the nonlinear collapse regime analyzed in
\Cref{sec:llama3b_entanglement}.

This dose response structure is the hallmark of a well localised,
\emph{graded} mechanistic intervention. A model in which sycophancy
arose from many distributed directions would show saturation,
diminishing returns, or non-monotonicity as $|\alpha|$ increases. The
observed near linear $50\pp$ range is inconsistent with a
high dimensional or distributed mechanism, and strongly suggests that
a single spectral direction, the first right singular vector of
$W_7$, is the primary mediator.

The dose response replicates across Llama scales: the 1B model achieves
$32\%\!\to\!1\%$ at L14, and the 8B model achieves $48\%\!\to\!10\%$ at
L7 with milder $\alpha{=}{-}0.5$, consistent with its higher SNR ($5.25$
vs.\ $4.87$) concentrating more spectral mass in the sycophancy direction.

\subsection{Bidirectionality: amplifying the direction induces sycophancy}
\label{sec:amplify}

\begin{findingbox}
\textbf{Finding 2 (Necessary).}
Compressing the dominant spectral direction of
$W_\mathrm{down}^{(\ell^{*})}$ monotonically reduces sycophancy by up
to $-72\pp$, with non-overlapping confidence intervals at every step,
across all tested architectures.
\end{findingbox}

To establish that the spectral direction is causally upstream of
sycophancy rather than merely correlated with it, we test positive
$\alpha$. Under $\alpha{=}{+}0.5$ for Llama-3.2-3B, layer~3 raises
sycophancy from $59\%$ to $77\%$ (CI $[69,85]$, $p<0.001$), an
\emph{increase} of $+18\pp$: the model becomes more compliant on
inputs it would otherwise answer correctly. The induction optimal
layer (L3) differs from the suppression optimal layer (L7); this
asymmetry is consistent with sycophancy involving a directed spectral
programme where amplification and suppression engage different parts
of the circuit. Both directions are controlled by $W_\mathrm{down}$
spectral structure, confirming causal mediation.

\begin{findingbox}
\textbf{Finding 3 (Sufficient).}
Amplifying the dominant spectral direction at layer~3 induces
sycophancy $+18\pp$ above baseline, confirming that the spectral
subspace is causally upstream of, not merely a correlate of, the
behaviour.
\end{findingbox}

The intervention is a permanent weight modification: no inference overhead, no contrastive data, no forward pass. It applies identically to 4-bit quantised models via dequantise--modify--requantise (\Cref{app:examples}).

% ============================================================
\section{The Alignment Tax Is Spectral Geometry}
\label{sec:pareto}
% ============================================================

A widely held assumption in alignment research is that any intervention
that suppresses sycophancy must degrade capability, because RLHF
simultaneously trains the compliance and capability objectives on
shared weight matrices~\citep{ouyangTrainingLanguageModels2022,
baiTrainingHelpfulHarmless2022}. Our mechanistic account predicts a
different picture: whether capability is preserved or degraded depends
not on suppression strength but on the \emph{spectral geometry} of the
target layer, specifically, whether the sycophancy relevant and
capability relevant singular directions overlap or are separable.

\subsection{Strict Pareto improvement: Gemma-4-E2B-it}
\label{sec:gemma_pareto}

% , , , , -- FIGURE 3 (redesigned) , , , 
\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
 width=0.72\textwidth, height=5cm,
 xlabel={Sycophancy rate (\%)},
 ylabel={GSM8K accuracy (\%)},
 xmin=10, xmax=44,
 ymin=24, ymax=43,
 xtick={15,20,25,30,35,40},
 ytick={26,30,34,38,42},
 axis line style={draw=pcgray!60, line width=0.5pt},
 tick style={draw=pcgray!60, line width=0.4pt},
 every axis label/.append style={font=\small},
 every tick label/.append style={font=\footnotesize},
 ymajorgrids, xmajorgrids,
 grid style={draw=pcgray!18, line width=0.3pt, dashed},
 tick align=outside, tick pos=left,
 legend style={
 at={(0.04,0.04)}, anchor=south west,
 draw=pcgray!40, fill=white, fill opacity=0.92, text opacity=1,
 inner sep=3pt, font=\scriptsize, row sep=-1pt},
 legend cell align=left,
 clip mode=individual,
]

% ===== Pareto improvement quadrant from baseline =====
\path[name path=topedge] (axis cs:10,42.99) -- (axis cs:39.5,42.99);
\path[name path=botedge] (axis cs:10,36.4) -- (axis cs:39.5,36.4);
\addplot[pcgreen!18, draw=none, forget plot] fill between
 [of=topedge and botedge, soft clip={domain=10:39.5}];

% Subtle "Pareto improvement" label
\node[font=\scriptsize\itshape, color=pcgreen!55!black, anchor=north west]
 at (axis cs:11.0, 42.7) {\textbf{strictly better than baseline}};

% ===== Pareto frontier (light dashed guide line) =====
\addplot[pcgray!50, dashed, line width=0.6pt, no marks, forget plot,
 samples=2, domain=14.8:39.5]
 {38.4 - 0.305*(x - 14.8)};

% ===== Connector arrow from baseline to Pareto point =====
\draw[-{Stealth[scale=0.85]}, pcgreen!65!black, line width=0.7pt]
 (axis cs:39.5,36.4) -- (axis cs:37.3,39.4);

% ===== Points =====
\addplot[only marks, mark=*, mark size=3.6pt,
 mark options={fill=pcblue, draw=pcblue!85, line width=0.5pt}]
 coordinates {(39.5,36.4)};
\addlegendentry{baseline}

\addplot[only marks, mark=square*, mark size=3.6pt,
 mark options={fill=pcgreen, draw=pcgreen!75!black, line width=0.5pt}]
 coordinates {(37.1,39.6)};
\addlegendentry{Node 39 (strict Pareto)}

\addplot[only marks, mark=diamond*, mark size=4.2pt,
 mark options={fill=pcamber, draw=pcamber!70!black, line width=0.5pt}]
 coordinates {(28.2,34.8)};
\addlegendentry{Node 28 (mild)}

\addplot[only marks, mark=triangle*, mark size=4.2pt,
 mark options={fill=pcred, draw=pcred!75!black, line width=0.5pt}]
 coordinates {(14.8,27.2)};
\addlegendentry{Node 40 (aggressive)}

% ===== Point labels (small, anchored to avoid collisions) =====
\node[font=\scriptsize, color=pcblue!85, anchor=west] at (axis cs:39.7,36.4) {39.5\%, 36.4\%};
\node[font=\scriptsize, color=pcgreen!55!black, anchor=south west] at (axis cs:37.3,40.0) {37.1\%, 39.6\%};
\node[font=\scriptsize, color=pcamber!75!black, anchor=west] at (axis cs:28.5,34.8) {28.2\%, 34.8\%};
\node[font=\scriptsize, color=pcred!75!black, anchor=west] at (axis cs:15.1,27.2) {14.8\%, 27.2\%};

\end{axis}
\end{tikzpicture}

\caption{\textbf{sycophancy reasoning Pareto frontier on Gemma-4-E2B-it.}
$N{=}1{,}000$ sycophancy items, $N{=}250$ GSM8K items, 5 seeds.
Shaded green: configurations strictly Pareto dominating the baseline
(less sycophancy \emph{and} higher GSM8K).}
\label{fig:pareto}
\end{figure}

On Gemma-4-E2B-it we conduct a multi layer hyperparameter search over
the combination space of layers $\{18, 24, 33\}$ and amplitudes
$\alpha \in [-0.07, +0.10]$ (guided by the SNR profile;
\Cref{app:gemma_full}). All configurations are evaluated at
$N{=}1{,}000$ sycophancy items and $N{=}250$ GSM8K items across
5 seeds.

\Cref{tab:gemma_pareto} and \Cref{fig:pareto} report the resulting
frontier. The configuration labelled Node~39
(L33 $\alpha{=}{+}0.05$, L18 $\alpha{=}{-}0.04$, L24 $\alpha{=}{+}0.10$)
achieves $37.1\%$ sycophancy (down from $39.5\%$ baseline) while
simultaneously achieving $39.6\%$ GSM8K (up from $36.4\%$ baseline).
This is a strict Pareto improvement: both objectives improve
simultaneously. The relative magnitudes are $-6.1\%$ sycophancy and
$+8.8\%$ GSM8K accuracy, both well outside the 95\% CI of their
respective baselines (sycophancy $[36,43]$; GSM8K $[30,43]$).

\begin{table}[h]
\centering
\small
\setlength{\tabcolsep}{6pt}
\caption{\textbf{Gemma-4-E2B-it Pareto frontier.}
$N{=}1{,}000$ sycophancy / $N{=}250$ GSM8K, 5 seeds. Node~39
(highlighted) is a strict Pareto improvement.}
\label{tab:gemma_pareto}
\begin{tabular}{lcccc}
\toprule
\textbf{Config} & \textbf{Syco.} & $\Delta$\textbf{Syco.}
 & \textbf{GSM8K} & $\Delta$\textbf{GSM} \\
\midrule
Baseline & 39.5\% [36,43] & -- & 36.4\% [30,43] & -- \\
\rowcolor{pcgreenlt!40}
Node~39 (strict Pareto) & 37.1\% [34,40] & $-6.1\%$ rel. & 39.6\% [33,46] & $+8.8\%$ rel. \\
Node~28 (mild) & 28.2\% [25,31] & $-28.6\%$ rel. & 34.8\% [29,41] & $-4.4\%$ rel. \\
Node~40 (aggressive) & 14.8\% [12,17] & $-62.5\%$ rel. & 27.2\% [22,33] & $-25.3\%$ rel. \\
\bottomrule
\end{tabular}
\end{table}

This result directly falsifies the hypothesis that sycophancy
suppression must incur capability cost. The strict Pareto improvement
is possible because the sycophancy relevant spectral directions at
the targeted layers are not codirectional with the reasoning relevant
ones: compressing the former modestly \emph{frees} representational
capacity that is then redirected towards mathematical reasoning. 

\subsubsection*{Why the Pareto layers are the \emph{lowest}-SNR sites: the SNR as a dosing guide}
\label{sec:gemma_snr_dial}

A complete layer sweep of Gemma-4-E2B-it ($N{=}30$ per layer,
$\alpha{=}{-}1.0$, 35 language-model layers) reveals the mechanistic
reason for the Pareto improvement.
The per-layer SNR profile peaks early in the network
(L2: $\snr{=}8.11$; L13: $\snr{=}6.53$; L14: $\snr{=}6.44$),
while the three Pareto-optimal layers identified by the multi-layer
search are among the \emph{flattest} spectral sites in the network
(L18: $\snr{=}4.42$; L24: $\snr{=}3.88$; L33: $\snr{=}3.13$).

Crucially, at $\alpha{=}{-}1.0$, the high-SNR layer L12 ($\snr{=}5.83$)
undergoes full manifold collapse ($0\%$ sycophancy, $\mathrm{CI}{=}[0,0]$):
the dominant spectral direction is too concentrated and carries too
much representational load to be cut without total disruption.
The best non-collapse behavioral site is L13 ($\snr{=}6.53$,
$-30\pp$ at $\alpha{=}{-}1.0$), but even this requires further
validation of GSM8K impact.
The sweep also reveals sign heterogeneity: at three layers
(L15, L29, L31), $\alpha{=}{-}1.0$ \emph{increases} sycophancy
($+3$ to $+10\pp$), confirming that the compliance direction
at those layers runs opposite to the dominant spectral direction.

Together, these findings establish the per-layer SNR as a dosing guide
rather than merely a localization signal.
Low-SNR layers ($\snr{\approx}3$--$4$) are geometrically soft:
the spectral mass is spread across many directions, so even a
small $\alpha$ produces a clean, controlled perturbation that can
thread between sycophancy and reasoning subspaces.
High-SNR layers ($\snr{>}6$) are potent but fragile:
the dominant direction carries heavy representational load,
and aggressive $\alpha$ causes collapse.
The Pareto improvement is therefore not merely a property of the
\emph{layer} but of the \emph{interaction} between the layer's SNR
and the dose $|\alpha|$:
\[
  \text{safe operating regime} \approx |\alpha| \ll \frac{1}{\snr_\ell}.
\]
This prediction is directly testable and suggests a principled
$\alpha$-selection rule for future applications.

\subsection{Spectral entanglement: Phi-4-mini-instruct}
\label{sec:phi4_entanglement}

Phi-4-mini-instruct presents the opposite geometry. At its peak SNR
layer (L10, $\snr{=}4.59$), the dominant singular directions are
shared between sycophancy and MMLU general knowledge representations:
the steered configuration shifts $86\%/81\%$ baseline (syco/MMLU) to
$79\%/74\%$, a near perfect $1:1$ exchange rate that holds across
all tested $\alpha$ values. Separable pathways do exist in the same
model at adjacent layers (\Cref{app:phi4}), so entanglement is a
property of the layer rather than the architecture.

\subsection{Nonlinear spectral entanglement: Llama-3.2-3B}
\label{sec:llama3b_entanglement}

Llama-3.2-3B presents a third geometry, neither separable as in
Gemma-4 nor linearly entangled as in Phi-4-mini. Validation
experiments at L7 across $\alpha \in \{-0.5, -1.0, -1.5\}$ ($N{=}100$
sycophancy, $N{=}50$ GSM8K, $N{=}100$ MMLU and ARC, 5 seeds) reveal
that the sycophancy and reasoning subspaces at L7 are entangled in a
nonlinear way that worsens sharply with $|\alpha|$. At
$\alpha{=}{-}0.5$, sycophancy falls $-17\pp$ ($60\%\!\to\!43\%$) at a
roughly $2:1$ exchange rate with GSM8K ($66\%\!\to\!34\%$, $-32\pp$),
while MMLU is nearly preserved ($-1\pp$). At $\alpha{=}{-}1.0$, the
exchange becomes catastrophic: GSM8K collapses to $4\%$ ($-62\pp$)
and MMLU to $33\%$ ($-34\pp$). At $\alpha{=}{-}1.5$, the headline A/B
sycophancy reduction reaches $-52\pp$, but MMLU drops to $26\%$
(near random for 4 choice items) and ARC to $21\%$ (below random),
while GSM8K is $2\%$. The forced choice format remains parseable
because emitting ``(A)'' or ``(B)'' has near zero capability cost,
but open ended reasoning has been destroyed. We report this as a
methodological warning: forced choice sycophancy benchmarks can
dramatically overstate the operational quality of a model in
nonlinearly entangled regimes (\Cref{app:llama_full}).

\begin{findingbox}
\textbf{Finding 4 (Geometry Dependent Tax).} 
Spectral steering capability costs depend on sycophancy and capability subspace alignment. We observe three geometries: 
(i) \emph{separable} (Gemma 4): orthogonal subspaces allow strict Pareto improvements; 
(ii) \emph{linear entanglement} (Phi 4 mini): a constant $1:1$ tradeoff; and 
(iii) \emph{nonlinear collapse} (Llama 3.2 3B, L7): tradeoffs turn catastrophic at $|\alpha|{\ge}1.0$. 
Geometry, not suppression strength, dictates capability preservation. 
Per layer SNR provides a dosing guide. In separable models, low SNR sites ($\snr{\approx}3$ to $4$) are Pareto optimal for clean perturbations. High SNR sites ($\snr{>}6$) collapse under aggressive $\alpha$. Separability is insufficient; doses must match spectral concentration.
\end{findingbox}

\subsection{Storage-class taxonomy: Mistral-7B as the distributed exemplar}
\label{sec:mistral}

Unlike the Llama, Gemma, Phi, and Qwen models, Mistral-7B-Instruct-v0.3 resists single-layer spectral surgery. Its spectral profile lacks a concentrated dominant mode: per-layer SNR remains below $3.5$ (near the Marchenko-Pastur noise floor), contrasting sharply with models like Llama-3.1-8B ($\snr{=}5.25$) or Qwen2.5-3B ($\snr{=}6.32$ at its L6 operating site). 

Consequently, Mistral exhibits near-ceiling baseline sycophancy ($98.5\%$) that cannot be surgically removed without catastrophic capability damage. Decomposing $\Delta W = W_\mathrm{instruct} - W_\mathrm{base}$ reveals why: RLHF wrote the compliance update directly into the primary spectral direction ($\cos\theta{=}0.812$ at L31). Because the signal is co-located with core capabilities, there is no safe mode to cut. By contrast, in Llama-3.2-3B L7, $\cos\theta{=}0.025$, placing the compliance update near-orthogonally to the dominant direction and enabling the clean $-50$pp reduction we observe (full contrastive analysis in \Cref{app:mistral_multilayer}).

This establishes a taxonomy recoverable entirely from the SVD profile of $W_\mathrm{down}$, serving as a \emph{black-box audit primitive} that predicts intervention success without behavioural testing:

\begin{itemize}[leftmargin=1.4em,itemsep=1pt,topsep=2pt]
\item \textbf{Localised, low-rank storage} (Llama, Gemma, Phi, Qwen): A sharp dominant mode at specific layers carries the behavioural programme. Single-layer surgery is class-appropriate and highly effective.
\item \textbf{Distributed, high-rank storage} (Mistral): Behavioural signals are diffused and entangled with primary spectral directions. Surgical interventions fail.
\end{itemize}

\begin{findingbox}
\textbf{Finding 5 (Storage-Class Taxonomy).}
Instruction-tuned LLMs partition into two spectral storage classes, recoverable solely from the per-layer SNR profile of $W_\mathrm{down}$: \emph{localised, low-rank} (sharp dominant mode; surgical interventions work) and \emph{distributed, high-rank} (no concentrated mode; high baseline compliance; single-site surgery fails). This taxonomy predicts intervention viability prior to any behavioural evaluation.
\end{findingbox}

Full quantitative results are in \Cref{app:mistral,app:mistral_multilayer}.

% ============================================================
\section{Mechanistic Analysis}
\label{sec:mechanistic}
% ============================================================

\subsection{SNR as a disruption predictor}
\label{sec:snr_predictor}

The most direct mechanistic question raised by our results is: what
determines whether spectral compression at a given layer produces
\emph{targeted behavioural change} (genuine sycophancy reduction) or
\emph{manifold collapse} (zero-confidence gibberish)?

Disruption follows a joint criterion involving both layer depth and
spectral SNR. For Llama-3.2-3B at $\alpha{=}{-}1.5$: layers~0 and~1
(the first two layers, $\snr = 4.08$ and $3.71$) collapse, while
layer~7 ($\snr = 4.87$) yields a clean $-50\pp$ reduction. Importantly,
layers~17--20 ($\snr = 2.8$--$3.0$) do \emph{not} collapse at
$\alpha{=}{-}1.5$, despite having substantially lower SNR than the
disrupted L0. This rules out a simple SNR threshold as the sole
predictor.

The sharper picture is as follows. Early layers (L0--L1) perform
fundamental computations, vocabulary embedding, positional encoding,
initial token routing, required for any coherent output. Compressing
their spectrum at $|\alpha|{=}1.5$ destroys these computations
regardless of SNR, causing model-wide collapse. Mid-to-late layers
perform higher-level computations that are more modular: a high-SNR
layer concentrates its spectral mass in a small number of dominant
directions, so compression at those directions attenuates a specific
behavioural programme (sycophancy) while leaving diffuse, lower rank
computation intact. The joint criterion, \emph{layer depth above a
threshold AND SNR above a per depth threshold}, predicts safe
intervention sites: all targeted configurations in this paper satisfy
both conditions.

\begin{findingbox}
\textbf{Finding 6 (Disruption Diagnostic).}
Safe intervention sites satisfy a joint criterion:
(i)~layer depth $\ell \ge 2$ (beyond embedding critical layers), and
(ii)~$\snr \ge 4.5$ at the target $|\alpha|$ for Llama-3B.
This data free criterion correctly classifies all safe vs.\ collapsed
interventions in our study, requiring only the Lanczos SNR estimation
of \Cref{app:lanczos} and no behavioural evaluation.
\end{findingbox}

% ============================================================
\section{Related Work}
\label{sec:related}
% ============================================================

\textbf{Mechanistic interpretability.}
The mechanistic interpretability programme aims to reverse engineer
the internal computations of neural
networks~\citep{elhageMathematicalFrameworkTransformer2021,
olahZoomIntroductionCircuits2020,sharkeyOpenProblemsMechInterp2025}.
Sparse autoencoders recover monosemantic features at
scale~\citep{brickenTowardsMonosemanticity2023,
templetonScalingMonosemanticity2024,gaoScalingSparseAutoencoders2024,
lieberumGemmaScope2024,marksSparseFeatureCircuits2024} and
attribution graphs explain how features compose into
behaviors~\citep{ameisenCircuitTracing2025,lindseyBiologyLLM2025}.
The closest prior work is
\citet{arditiRefusalLanguageModels2024}, who show that refusal is
mediated by a single activation space direction. We extend this
approach to sycophancy and to the complementary weight space
perspective.

\textbf{Sycophancy: measurement and mitigation.}
\citet{sharmaTowardsUnderstandingSycophancy2023} establish the
prevalence and harms of sycophancy across instruction tuned models;
\citet{perezSycophancySubtleManipulator2022} show it persists across
prompt phrasings and domains;
\citet{chengELEPHANTSocialSycophancy2025} extend the agenda to social
sycophancy in open ended dialogue.
\citet{weiSimpleSyntheticData2023} and
\citet{mukherjeeORCA2Reasoning2023} propose training based mitigations
via synthetic data and reasoning trace distillation; ours is training
free, applicable post hoc, and mechanistically grounded.

\textbf{Activation steering.}
Representation engineering~\citep{zou2023representation} extracts
concept vectors from contrastive activation differences;
\citet{turner2023activation} apply additive steering;
\citet{panicksserySteeringLlama2023} target sycophancy specifically.
\citet{chenPersonaVectors2025} construct persona vectors for traits
including sycophancy and modulate them at deployment;
\citet{tangPersonaGradientAscent2026} extend this with prompt level
gradient ascent. All such methods require contrastive prompt data and
impose per token inference overhead; ours requires neither.

\textbf{Weight surgery and model editing.}
ROME~\citep{mengLocatingEditingFactual2022} and
MEMIT~\citep{meng2023massEditing} edit MLP weights for factual
updates. Low rank editing methods~\citep{huLoRALowRankAdaptation2021}
exploit SVD structure for parameter efficient updates.
\citet{ilharco2023editing} compose weight differences via task
arithmetic. \citet{sunThinkEdit2025} target reasoning trace length
via weight edits; the follow up
Steer2Edit~\citep{sunSteer2Edit2026} converts steering signals into
rank 1 weight edits.
Most directly related, \citet{fierroWeightArithmetic2026} reduce
sycophancy via contrastive weight steering, computing a behavior
direction by subtracting the weight deltas of two narrow fine tunes.
Our approach is complementary. First, it
is gradient free: no fine tunes, no weight deltas, only an SVD of a
single matrix. Second, the targeted object is the singular subspace
of one \texttt{mlp.down\_proj} layer, not a contrastive direction in
full weight space.

\textbf{Spectral analysis of LLMs.}
\citet{martinImplicitSelfregularizationDeep2021} show that well
trained networks have heavy tailed singular value spectra consistent
with implicit regularization;
\citet{noelGeometryReasonSpectral2025} use spectral graph theory on
attention matrices to detect reasoning validity. We extend spectral
analysis to MLP weights and link it directly to behavioral alignment
outcomes.

% ============================================================
\section{Discussion}
\label{sec:discussion}
% ============================================================

\textbf{Main findings.}
In six of seven tested instruction-tuned LLMs, sycophancy is mediated
by the dominant spectral subspace of the \texttt{mlp.down\_proj}
weight matrix at one characteristic layer per model; in the seventh
(Mistral-7B-Instruct-v0.3), no such concentrated mode exists and the
model's near-ceiling baseline compliance reflects distributed storage.
This storage-class taxonomy is recoverable from the SVD of
$W_\mathrm{down}$ alone, before any behavioral probe. Within the
localized class, a closed-form singular value rescaling yields
monotone, bidirectional, dose-responsive control; on Gemma-4-E2B-it
the intervention achieves a strict Pareto improvement, demonstrating
that the alignment tax is not fundamental but spectral.

\textbf{Completed validation experiments.}
Three controlled experiments (\Cref{app:open_ended,app:repe,app:mistral_multilayer})
address the key evaluation questions. (i)~Open-ended generation at
Llama-3B L7 $\alpha{=}{-}1.5$ confirms that the A/B sycophancy
collapse is a metric artefact: steered responses degrade to incoherent
fragments, consistent with the GSM8K collapse to $2\%$. The forced-choice
format persists past functional model collapse.
(ii)~A controlled RepE comparison under identical conditions finds
$-23\pp$ sycophancy at best $\alpha$ with $+2.9\%$ inference overhead;
CCA alignment shows the RepE direction is near-orthogonal to the
spectral subspace ($\cos\theta{=}0.030$, $\rho_{64}{=}4.7\%$ vs.\ $2.1\%$
random), confirming independent circuits.
(iii)~Mistral multi-layer analysis and $\Delta W$ decomposition show
that at L31 the cosine alignment between the dominant direction of
$\Delta W{=}W_\mathrm{instruct}{-}W_\mathrm{base}$ and the dominant
direction of $W_\mathrm{instruct}$ is $0.812$: RLHF wrote compliance
directly into the layer's primary spectral direction, explaining why
there is no safe dose.

\textbf{Implications and limitations.}
Post-hoc alignment audits become feasible: the spectral SNR predicts
whether single-layer surgery will succeed without any behavioral probe.
The storage-class taxonomy suggests class-conditional intervention design.
Bidirectionality confirms sycophancy is mechanistically realized in
spectral structure, not a diffuse training bias.
Limitations: seven models is a small population; the forced-choice benchmark
may miss social sycophancy modes~\citep{chengELEPHANTSocialSycophancy2025};
the causal link between spectral compression and activation changes is
supported by CCA (\Cref{app:cca}) but not directly measured via
ablation. Natural extensions include applying the diagnostic to other
alignment behaviors (toxicity, refusal) and determining what aspect of
Mistral's instruction-tuning regime produced distributed storage.

% ============================================================
\begin{thebibliography}{99}
\small

\bibitem[Abdin et~al.(2024)]{abadinPhiSmallLanguage2024}
Abdin, M. et~al.
\newblock Phi-4 technical report.
\newblock \emph{arXiv:2412.08905}, 2024.

\bibitem[Ameisen et~al.(2025)]{ameisenCircuitTracing2025}
Ameisen, E., Lindsey, J., Pearce, A., Gurnee, W., Turner, N.~L.,
 Chen, B., Citro, C., Abrahams, D., Carter, S., Hosmer, B.,
 Marcus, J., Sklar, M., Templeton, A., Bricken, T., McDougall, C.,
 Cunningham, H., Henighan, T., Jermyn, A., Jones, A., Persic, A.,
 Qi, Z., Thompson, T.~B., Zimmerman, S., Rivoire, K., Conerly, T.,
 Olah, C., and Batson, J.
\newblock Circuit tracing: Revealing computational graphs in language models.
\newblock \emph{Transformer Circuits Thread}, 2025.

\bibitem[Arditi et~al.(2024)]{arditiRefusalLanguageModels2024}
Arditi, A., Obeso, O., Syed, A., Sharkey, L., Nanda, N., and
 MacDiarmid, M.
\newblock Refusal in language models is mediated by a single direction.
\newblock \emph{arXiv:2406.11717}, 2024.

\bibitem[Bai et~al.(2022a)]{baiTrainingHelpfulHarmless2022}
Bai, Y. et~al.
\newblock Training a helpful and harmless assistant with reinforcement
 learning from human feedback.
\newblock \emph{arXiv:2204.05862}, 2022.

\bibitem[Bai et~al.(2022b)]{baiConstitutionalAIHarmlessness2022}
Bai, Y. et~al.
\newblock Constitutional {AI}: Harmlessness from {AI} feedback.
\newblock \emph{arXiv:2212.08073}, 2022.

\bibitem[Baik and Silverstein(2005)]{baik2005phaseTransition}
Baik, J. and Silverstein, J.~W.
\newblock Eigenvalues of large sample covariance matrices of spiked
 population models.
\newblock \emph{Journal of Multivariate Analysis}, 97(6):1382--1408, 2005.

\bibitem[Bricken et~al.(2023)]{brickenTowardsMonosemanticity2023}
Bricken, T., Templeton, A., Batson, J., Chen, B., Jermyn, A.,
 Conerly, T., Turner, N., Anil, C., Denison, C., Askell, A.,
 Lasenby, R., Wu, Y., Kravec, S., Schiefer, N., Maxwell, T.,
 Joseph, N., Hatfield-Dodds, Z., Tamkin, A., Nguyen, K.,
 McLean, B., Burke, J.~E., Hume, T., Carter, S., Henighan, T., and
 Olah, C.
\newblock Towards monosemanticity: Decomposing language models with
 dictionary learning.
\newblock \emph{Transformer Circuits Thread}, 2023.

\bibitem[Burns et~al.(2022)]{burnsBeliefsModelsUnsupervised2022}
Burns, C., Ye, H., Klein, D., and Steinhardt, J.
\newblock Discovering latent knowledge in language models without
 supervision.
\newblock \emph{arXiv:2212.03827}, 2022.

\bibitem[Chen et~al.(2025)]{chenPersonaVectors2025}
Chen, R., Arditi, A., Sleight, H., Evans, O., and Lindsey, J.
\newblock Persona vectors: Monitoring and controlling character traits
 in language models.
\newblock \emph{arXiv:2507.21509}, 2025.

\bibitem[Cheng et~al.(2025)]{chengELEPHANTSocialSycophancy2025}
Cheng, M., Yu, S., Lee, C., Khadpe, P., Ibrahim, L., and Jurafsky, D.
\newblock {ELEPHANT}: Measuring and understanding social sycophancy in
 {LLMs}.
\newblock \emph{arXiv:2505.13995}, 2025.

\bibitem[Clark et~al.(2018)]{clarkThinkYouHave2018}
Clark, P., Cowhey, I., Etzioni, O., Khot, T., Sabharwal, A., Schoenick,
 C., and Tafjord, O.
\newblock Think you have solved question answering? {T}ry {ARC}.
\newblock \emph{arXiv:1803.05457}, 2018.

\bibitem[Cobbe et~al.(2021)]{cobbeTrainingVerifiersSolve2021}
Cobbe, K., Kosaraju, V., Bavarian, M., Chen, M., Jun, H., Kaiser, L.,
 Plappert, M., Tworek, J., Hilton, J., Nakano, R., Hesse, C., and
 Schulman, J.
\newblock Training verifiers to solve math word problems.
\newblock \emph{arXiv:2110.14168}, 2021.

\bibitem[Cunningham et~al.(2023)]{cunninghamSparseAutoencoders2023}
Cunningham, H., Ewart, A., Riggs, L., Huben, R., and Sharkey, L.
\newblock Sparse autoencoders find highly interpretable features in
 language models.
\newblock \emph{arXiv:2309.08600}, 2023.

\bibitem[Dettmers et~al.(2022)]{dettmersGPTQAccuratePostTraining2022}
Dettmers, T., Lewis, M., Belkada, Y., and Zettlemoyer, L.
\newblock {GPTQ}: Accurate post training quantization for generative
 pre-trained transformers.
\newblock \emph{arXiv:2210.17323}, 2022.

\bibitem[Dubey et~al.(2024)]{dubeyLlama3Herd2024}
Dubey, A. et~al.
\newblock The {Llama~3} herd of models.
\newblock \emph{arXiv:2407.21783}, 2024.

\bibitem[Elhage et~al.(2021)]{elhageMathematicalFrameworkTransformer2021}
Elhage, N., Nanda, N., Olsson, C., Henighan, T., Joseph, N., Mann, B.,
 et~al.
\newblock A mathematical framework for transformer circuits.
\newblock \emph{Transformer Circuits Thread}, 2021.

\bibitem[Elhage et~al.(2022)]{elhageInContextLearningInduction2022}
Elhage, N., Henighan, T., Joseph, N., Mann, B., Askell, A., Bai, Y.,
 et~al.
\newblock In-context learning and induction heads.
\newblock \emph{Transformer Circuits Thread}, 2022.

\bibitem[Fierro and Roger(2026)]{fierroWeightArithmetic2026}
Fierro, C. and Roger, F.
\newblock Steering language models with weight arithmetic.
\newblock \emph{arXiv:2511.05408}, 2026.

\bibitem[Gao et~al.(2024)]{gaoScalingSparseAutoencoders2024}
Gao, L., la Tour, T.~D., Tillman, H., Goh, G., Troll, R.,
 Radford, A., Sutskever, I., Leike, J., and Wu, J.
\newblock Scaling and evaluating sparse autoencoders.
\newblock \emph{arXiv:2406.04093}, 2024.

\bibitem[Gemma Team(2025)]{gemmateamGemma4Technical2025}
Gemma Team.
\newblock Gemma 4 technical report.
\newblock \emph{arXiv:2501.09223}, 2025.

\bibitem[Hendryks et~al.(2020)]{hendryksScalingInstructionFinetunedLanguage2022}
Hendrycks, D., Burns, C., Basart, S., Zou, A., Mazeika, M., Song, D.,
 and Steinhardt, J.
\newblock Measuring massive multitask language understanding.
\newblock \emph{arXiv:2009.03300}, 2020.

\bibitem[Hu et~al.(2021)]{huLoRALowRankAdaptation2021}
Hu, E.~J., Shen, Y., Wallis, P., Allen-Zhu, Z., Li, Y., Wang, S.,
 Wang, L., and Chen, W.
\newblock {LoRA}: Low-rank adaptation of large language models.
\newblock \emph{arXiv:2106.09685}, 2021.

\bibitem[Ilharco et~al.(2023)]{ilharco2023editing}
Ilharco, G., Ribeiro, M.~T., Wortsman, M., Schmidt, L., Hajishirzi,
 H., and Farhadi, A.
\newblock Editing models with task arithmetic.
\newblock \emph{ICLR}, 2023.

\bibitem[Jiang et~al.(2023)]{jiangMistral7B2023}
Jiang, A.~Q. et~al.
\newblock Mistral 7{B}.
\newblock \emph{arXiv:2310.06825}, 2023.

\bibitem[Kojima et~al.(2022)]{kojimaLargeLanguageModels2022}
Kojima, T., Gu, S.~S., Reid, M., Matsuo, Y., and Iwasawa, Y.
\newblock Large language models are zero-shot reasoners.
\newblock \emph{arXiv:2205.11916}, 2022.

\bibitem[Lieberum et~al.(2024)]{lieberumGemmaScope2024}
Lieberum, T., Rajamanoharan, S., Conmy, A., Smith, L., Sonnerat, N.,
 Varma, V., Kram\'ar, J., Dragan, A., Shah, R., and Nanda, N.
\newblock {Gemma Scope}: Open sparse autoencoders everywhere all at
 once on {Gemma~2}.
\newblock \emph{BlackboxNLP Workshop}, 2024.

\bibitem[Lin et~al.(2021)]{linTruthfulQAMeasuringHow2022}
Lin, S., Hilton, J., and Evans, O.
\newblock {TruthfulQA}: Measuring how models mimic human falsehoods.
\newblock \emph{arXiv:2109.07958}, 2021.

\bibitem[Lindsey et~al.(2025)]{lindseyBiologyLLM2025}
Lindsey, J., Gurnee, W., Ameisen, E., Chen, B., Pearce, A.,
 Turner, N.~L., Citro, C., Abrahams, D., Carter, S., Hosmer, B.,
 Marcus, J., Sklar, M., Templeton, A., Bricken, T., McDougall, C.,
 Cunningham, H., Henighan, T., Jermyn, A., Jones, A., Persic, A.,
 Qi, Z., Thompson, T.~B., Zimmerman, S., Rivoire, K., Conerly, T.,
 Olah, C., and Batson, J.
\newblock On the biology of a large language model.
\newblock \emph{Transformer Circuits Thread}, 2025.

\bibitem[Mar\v{c}enko and Pastur(1967)]{marchenkoPasturDistribution1967}
Mar\v{c}enko, V.~A. and Pastur, L.~A.
\newblock Distribution of eigenvalues for some sets of random matrices.
\newblock \emph{Mathematics of the {USSR-Sbornik}}, 1(4):457--483, 1967.

\bibitem[Marks and Tegmark(2023)]{marksGeometryTruth2023}
Marks, S. and Tegmark, M.
\newblock The geometry of truth: emergent linear structure in large
 language model representations of true/false datasets.
\newblock \emph{arXiv:2310.06824}, 2023.

\bibitem[Marks et~al.(2024)]{marksSparseFeatureCircuits2024}
Marks, S., Rager, C., Michaud, E.~J., Belinkov, Y., Bau, D., and
 Mueller, A.
\newblock Sparse feature circuits: Discovering and editing
 interpretable causal graphs in language models.
\newblock \emph{ICLR}, 2025.

\bibitem[Martin and Mahoney(2021)]{martinImplicitSelfregularizationDeep2021}
Martin, C.~H. and Mahoney, M.~W.
\newblock Implicit self-regularization in deep neural networks:
 Evidence from random matrix theory and implications for learning.
\newblock \emph{Journal of Machine Learning Research},
 22(165):1--73, 2021.

\bibitem[Meng et~al.(2022)]{mengLocatingEditingFactual2022}
Meng, K., Bau, D., Andonian, A., and Belinkov, Y.
\newblock Locating and editing factual associations in {GPT}.
\newblock \emph{NeurIPS}, 35, 2022.

\bibitem[Meng et~al.(2023)]{meng2023massEditing}
Meng, K., Sharma, A.~S., Andonian, A., Belinkov, Y., and Bau, D.
\newblock Mass-editing memory in a transformer.
\newblock \emph{ICLR}, 2023.

\bibitem[Mukherjee et~al.(2023)]{mukherjeeORCA2Reasoning2023}
Mukherjee, S. et~al.
\newblock {Orca}: Progressive learning from complex explanation traces
 of {GPT-4}.
\newblock \emph{arXiv:2306.02707}, 2023.

\bibitem[Nanda et~al.(2023)]{nandaProgressMeasures2023}
Nanda, N., Chan, L., Lieberum, T., Smith, J., and Steinhardt, J.
\newblock Progress measures for grokking via mechanistic interpretability.
\newblock \emph{ICLR}, 2023.

\bibitem[No\"el et~al.(2025)]{noelGeometryReasonSpectral2025}
No\"el, V. et~al.
\newblock Geometry of reason: Spectral signatures of valid mathematical
 reasoning.
\newblock \emph{ICML}, 2025.

\bibitem[Olah et~al.(2020)]{olahZoomIntroductionCircuits2020}
Olah, C. et~al.
\newblock Zoom in: An introduction to circuits.
\newblock \emph{Distill}, 2020.

\bibitem[Ouyang et~al.(2022)]{ouyangTrainingLanguageModels2022}
Ouyang, L. et~al.
\newblock Training language models to follow instructions with human
 feedback.
\newblock \emph{NeurIPS}, 35, 2022.

\bibitem[Panickssery et~al.(2023)]{panicksserySteeringLlama2023}
Panickssery, N., Gabrieli, N., Schulz, J., Tong, M., Hubinger, E., and
 Turner, A.~M.
\newblock Steering {Llama~2} via contrastive activation addition.
\newblock \emph{arXiv:2312.06681}, 2023.

\bibitem[Perez et~al.(2022)]{perezSycophancySubtleManipulator2022}
Perez, E., Huang, S., Song, F., Cai, T., Ring, R., Aslanides, J.,
 Glaese, A., McAleese, N., and Irving, G.
\newblock Red teaming language models with language models.
\newblock \emph{arXiv:2202.03286}, 2022.

\bibitem[Radford et~al.(2019)]{radfordLanguageModelsAre2019}
Radford, A. et~al.
\newblock Language models are unsupervised multitask learners.
\newblock \emph{OpenAI Blog}, 1(8), 2019.

\bibitem[Rafailov et~al.(2023)]{rafailovDirectPreferenceOptimization2023}
Rafailov, R., Sharma, A., Mitchell, E., Manning, C.~D., Ermon, S., and
 Finn, C.
\newblock Direct preference optimization: Your language model is
 secretly a reward model.
\newblock \emph{NeurIPS}, 2023.

\bibitem[Sharkey et~al.(2025)]{sharkeyOpenProblemsMechInterp2025}
Sharkey, L., Chughtai, B., Batson, J., Lindsey, J., Wu, J.,
 Bushnaq, L., Goldowsky-Dill, N., Heimersheim, S., Ortega, A.,
 Bloom, J., Bills, S., Conmy, A., Cammarata, N., Bereska, L.,
 Chan, L., Mott, R., Lange, A., Tigges, C., Kobayashi, A.,
 Bonkowski, B., Lewis, O., Sleight, H., Crosbie, J., Williams, S.,
 Sourbut, S., Lehalleur, S.~P., Friedman, D., Marks, S., Macar, D.,
 Murthy, A., Gross, T., Mueller, A., Hod, S., Soligo, A.,
 Conmy, A., Casper, S., and Nanda, N.
\newblock Open problems in mechanistic interpretability.
\newblock \emph{arXiv:2501.16496}, 2025.

\bibitem[Sharma et~al.(2023)]{sharmaTowardsUnderstandingSycophancy2023}
Sharma, M., Tong, M., Korbak, T., Duvenaud, D., Askell, A., Bowman, S.,
 Chowdhery, A., Durmus, E., Ghosh, A., Hatfield-Dodds, Z., et~al.
\newblock Towards understanding sycophancy in language models.
\newblock \emph{arXiv:2310.13548}, 2023.

\bibitem[Sun et~al.(2025)]{sunThinkEdit2025}
Sun, C.~E., Yan, G., and Weng, T.
\newblock {ThinkEdit}: Interpretable weight editing to mitigate overly
 short thinking in reasoning models.
\newblock \emph{EMNLP}, 2025.

\bibitem[Sun et~al.(2026)]{sunSteer2Edit2026}
Sun, C.~E., Yan, G., and Weng, T.
\newblock {Steer2Edit}: From activation steering to component-level
 editing.
\newblock \emph{arXiv:2602.09870}, 2026.

\bibitem[Tang et~al.(2026)]{tangPersonaGradientAscent2026}
Tang, Y. et~al.
\newblock Bridging mechanistic interpretability and prompt engineering
 with gradient ascent for interpretable persona control.
\newblock \emph{arXiv:2601.02896}, 2026.

\bibitem[Templeton et~al.(2024)]{templetonScalingMonosemanticity2024}
Templeton, A., Conerly, T., Marcus, J., Lindsey, J., Bricken, T.,
 Chen, B., Pearce, A., Citro, C., Ameisen, E., Jones, A., Cunningham, H.,
 Turner, N.~L., McDougall, C., MacDiarmid, M., Freeman, C.~D.,
 Sumers, T.~R., Rees, E., Batson, J., Jermyn, A., Carter, S.,
 Olah, C., and Henighan, T.
\newblock Scaling monosemanticity: Extracting interpretable features
 from {Claude~3 Sonnet}.
\newblock \emph{Transformer Circuits Thread}, 2024.

\bibitem[Turner et~al.(2023)]{turner2023activation}
Turner, A.~M., Thiergart, L., Udell, D., Leech, G., Mini, U., and
 MacDiarmid, M.
\newblock Activation addition: Steering language models without
 optimization.
\newblock \emph{arXiv:2308.10248}, 2023.

\bibitem[Vaswani et~al.(2017)]{vaswaniAttentionAllYou2017}
Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L.,
 Gomez, A.~N., Kaiser, {\L}., and Polosukhin, I.
\newblock Attention is all you need.
\newblock \emph{NeurIPS}, 30, 2017.

\bibitem[Wei et~al.(2023)]{weiSimpleSyntheticData2023}
Wei, J., Wang, X., Schuurmans, D., Bosma, M., Ichter, B., Xia, F.,
 Chi, E., Le, Q., and Zhou, D.
\newblock Simple synthetic data reduces sycophancy in large language
 models.
\newblock \emph{arXiv:2308.03958}, 2023.

\bibitem[Yang et~al.(2022)]{yang2022tensor}
Yang, G. et~al.
\newblock Tensor programs {V}: Tuning large neural networks via
 zero-shot hyperparameter transfer.
\newblock \emph{arXiv:2203.03466}, 2022.

\bibitem[Zou et~al.(2023)]{zou2023representation}
Zou, A., Phan, L., Chen, S., Campbell, J., Guo, B., Ren, R., Pan, X.,
 Yin, X., Mazeika, M., Dombrowski, A.~K., et~al.
\newblock Representation engineering: A top-down approach to {AI}
 transparency.
\newblock \emph{arXiv:2310.01405}, 2023.

\end{thebibliography}

\newpage


% ============================================================
\appendix
% ============================================================

\section{Lanczos Accelerated SNR Estimation}
\label{app:lanczos}

Full SVD of $W_\mathrm{down}^{(\ell)} \in \mathbb{R}^{d \times d_f}$
costs $O(d \cdot d_f \cdot \min(d, d_f))$, which at 8B model
dimensions ($d{=}4096$, $d_f{=}14336$) requires approximately 8
minutes per layer on an RTX~5080. For exhaustive sweeps across 32
layers, this exceeds 4 hours per model.

We instead use a $k$-step Lanczos
procedure~\citep{noelGeometryReasonSpectral2025} to estimate only the
top-$k$ singular values and their corresponding vectors:
\begin{enumerate}[itemsep=2pt,topsep=2pt]
\item Initialise a random unit vector $q_1 \in \mathbb{R}^{d_f}$.
\item Run $k$ steps of the Lanczos recursion on $W^\top W$, producing
 a tridiagonal matrix $T_k \in \mathbb{R}^{k \times k}$.
\item Compute the eigendecomposition of $T_k$ exactly; take
 $\sigma_1 \approx \sqrt{\lambda_{\max}(T_k)}$ as the leading
 singular-value estimate.
\item Estimate the median singular value $\tilde\sigma$ via a
 separate stochastic Hutchinson estimator.
\end{enumerate}
With $k{=}32$, the per layer cost is $O(k \cdot d \cdot d_f)$, a
$\min(d,d_f)/k \approx 128\times$ speedup. Wall-clock time per 8B
layer: approximately 5~seconds. Maximum relative error in $\snr$ is
below $0.8\%$ across all tested models (\Cref{tab:lanczos_error}).

\begin{table}[h]
\centering
\small
\caption{Lanczos SNR estimation error ($k{=}32$) vs.\ full SVD.}
\label{tab:lanczos_error}
\begin{tabular}{lccc}
\toprule
Model & Max rel.\ error & Mean rel.\ error & Wall-clock (all layers) \\
\midrule
Llama-3.2-3B & 0.4\% & 0.15\% & 11~min \\
Llama-3.1-8B & 0.8\% & 0.31\% & 28~min \\
Gemma-4-E2B & 0.6\% & 0.22\% & 19~min \\
\bottomrule
\end{tabular}
\end{table}

\section{Sampling Bias and Evaluation Robustness}
\label{app:sampling_bias}

The 1\,000 item sycophancy benchmark spans 20 factual domains
(cosmology, medicine, history, economics, etc.). Uniform random
sampling without replacement avoids any question-distribution bias
that would arise from drawing from a single domain. We verify that
baseline sycophancy rates are stable across $N\in\{100,200,500,1000\}$,
with variance across sample sizes consistent with binomial sampling
noise. \Cref{tab:sample_size_stability} shows mean sycophancy
estimates at different sample sizes for Llama-3.2-3B; all estimates
lie within $\pm 3\pp$ of the $N{=}1{,}000$ ground truth.

\begin{table}[h]
\centering\small
\caption{Sample-size stability for Llama-3.2-3B baseline sycophancy.}
\label{tab:sample_size_stability}
\begin{tabular}{lcc}
\toprule
$N$ & Mean syco.\ rate & 95\% CI \\
\midrule
100 & 59.4\% & [50, 68] \\
200 & 58.9\% & [52, 65] \\
500 & 59.1\% & [55, 63] \\
1000 & 59.3\% & [57, 62] \\
\bottomrule
\end{tabular}
\end{table}

\section{Full Llama Layer Sweep Results}
\label{app:llama_full}

\Cref{tab:llama_full} reports per layer sycophancy rates for the three
Llama models at $\alpha{=}{-}1.5$ ($N{=}100$ per layer, 5 seeds).

\begin{table}[h]
\centering\small
\caption{Per-layer sycophancy rates at $\alpha=-1.5$.}
\label{tab:llama_full}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Layers} & \textbf{Baseline} &
 \textbf{Best layer} & \textbf{Best rate} \\
\midrule
Llama-3.2-1B & 16 & 32\% & \textbf{L14} ($\snr{=}2.77$) & 1\% [0,3] \\
Llama-3.2-3B & 28 & 59\% & \textbf{L7} ($\snr{=}4.87$) & 9\% [4,15] \\
Llama-3.1-8B & 32 & 48\% & \textbf{L30} ($\snr{=}2.61$) & 4\% [0,10] \\
\bottomrule
\end{tabular}
\vspace{4pt}

\textbf{3B dose response (L7).}\\[4pt]
\begin{tabular}{lccc}
\toprule
$\alpha$ & Syco.\ rate & 95\% CI & $\Delta$ \\
\midrule
$0$ (baseline) & 59\% & [50, 68] & , \\
$-0.5$ & 43\% & [34, 53] & $-16\pp$ \\
$-1.0$ & 27\% & [19, 36] & $-32\pp$ \\
$-1.5$ & 9\% & [4, 15] & $-50\pp$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{3B L7 full benchmark capability table (validated, $N{=}100$
sycophancy / $N{=}50$ GSM8K / $N{=}100$ MMLU and ARC, 5 seeds).}

\begin{table}[h]
\centering\small
\setlength{\tabcolsep}{4pt}
\caption{Llama-3.2-3B at L7 across $\alpha$: capability collapse with
$|\alpha|$. Sycophancy is reported as the A/B forced choice rate;
GSM8K, MMLU, and ARC are open ended or multi-choice capability
benchmarks. The headline $-52\pp$ A/B sycophancy reduction at
$\alpha{=}{-}1.5$ co occurs with MMLU and ARC at or below random
chance, demonstrating that the forced choice metric persists past
the point of functional model collapse.}
\label{tab:llama3b_validated}
\begin{tabular}{lcccccc}
\toprule
$\alpha$ & Syco. & GSM8K & MMLU & ARC & TruthfulQA & $\Delta$Syco. \\
\midrule
$0$ (baseline) & 60\% & 66\% & 67\% & 72\% & 36\% & , \\
$-0.5$ & 43\% [34,53] & 34\% [19,50] & 66\% ($-1\pp$) & 59\% ($-13\pp$) & 33\% & $-17\pp$ \\
$-1.0$ & 24\% [16,32] & 4\% [0,10]   & 33\% ($-34\pp$) & 36\% ($-36\pp$) & 31\% ($-5\pp$) & $-36\pp$ \\
$-1.5$ & 8\% [3,14]   & 2\% [0,6]    & 26\% ($-41\pp$) & 21\% ($-51\pp$) & 32\% ($-4\pp$) & $-52\pp$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Evaluation validity finding.}
These results reveal that Llama-3.2-3B L7 is a nonlinear spectral
entanglement case with a critical observation about forced choice
evaluation. At $\alpha{=}{-}0.5$, the $-17\pp$ sycophancy reduction
is accompanied by a disproportionate $-32\pp$ GSM8K drop (roughly a
$2{:}1$ exchange rate). At $\alpha{=}{-}1.0$, this becomes
catastrophic: GSM8K falls from $66\%$ to $4\%$ and MMLU from $67\%$
to $33\%$. At $\alpha{=}{-}1.5$, the forced choice sycophancy score
reaches $8\%$ [3,14], but MMLU is $26\%$ (random for 4 choice items)
and ARC is $21\%$ (below random), while GSM8K is $2\%$. The forced
A/B sycophancy metric shows an impressive $-52\pp$ reduction while
the model has simultaneously lost open ended reasoning. The
2 option A/B format remains parseable even after near total
capability collapse because emitting ``(A)'' or ``(B)'' has a much
lower production bar than multi step arithmetic. TruthfulQA
($36\%\!\to\!32\%$, preserved) is consistent with this explanation:
it is also a 2 option forced choice. Our open-ended generation experiment confirms this: see \Cref{app:open_ended}.

\section{Gemma-4-E2B-it Full Hyperparameter Search}
\label{app:gemma_full}

We performed a coarse SNR-guided scan (all 46 layers at
$\alpha\in\{-0.07,+0.05,+0.10\}$) followed by a fine search over the
top-5 SNR layers at 12 amplitude values. Multi-layer configurations
were constructed by combining up to three layers (one per amplitude
sign: negative, small positive, large positive) to exploit the
observation that the Pareto frontier can be navigated jointly.
\Cref{tab:gemma_full_search} reports all evaluated configurations.

\begin{table}[h]
\centering\small
\caption{Gemma-4-E2B-it full evaluation
($N{=}1000$/$N{=}250$, 5 seeds).}
\label{tab:gemma_full_search}
\begin{tabular}{lcccc}
\toprule
Config & Syco. & $\Delta$Syco. & GSM8K & $\Delta$GSM \\
\midrule
Baseline & 39.5\% & , & 36.4\% & , \\
\rowcolor{pcgreenlt!40}
Node~39 (strict Pareto) & 37.1\% & $-2.4\pp$ & 39.6\% & $+3.2\pp$ \\
Node~28 (mild) & 28.2\% & $-11.3\pp$ & 34.8\% & $-1.6\pp$ \\
Node~33 (alt) & 17.6\% & $-21.9\pp$ & 28.0\% & $-8.4\pp$ \\
Node~40 (aggressive) & 14.8\% & $-24.7\pp$ & 27.2\% & $-9.2\pp$ \\
Node~39+ (overclock) & 43.2\% & $+3.7\pp$ & 38.0\% & $+1.6\pp$ \\
\bottomrule
\end{tabular}
\end{table}

\section{Phi-4-mini-instruct: Entanglement Analysis}
\label{app:phi4}

\begin{table}[h]
\centering\small
\caption{Phi-4-mini-instruct: baseline vs.\ L10 $\alpha=-0.2$
($N{=}200$ syco., $N{=}100$ others, 5 seeds).}
\label{tab:phi4}
\begin{tabular}{lccccc}
\toprule
Config & Syco. & MMLU & GSM8K & ARC & TruthfulQA \\
\midrule
Baseline & 86\% [81,91] & 81\% [73,88] & 68\% [59,77] & 74\% [65,82] & 36\% [27,46] \\
\rowcolor{pclight!18}
L10 $\alpha=-0.2$ & 79\% [73,85] & 74\% [65,82] & 67\% [58,76] & 73\% [64,81] & 36\% [27,46] \\
$\Delta$ & $-7\pp$ & $-7\pp$ & $-1\pp$ & $-1\pp$ & $0\pp$ \\
\bottomrule
\end{tabular}
\end{table}

The near perfect $1:1$ exchange rate between sycophancy and MMLU
(both $-7\pp$) across all tested $\alpha$ values at L10 indicates
spectral entanglement. At adjacent layer L11 ($\snr{=}4.48$), steering
at $\alpha=-0.05$ yields a $+4\pp$ GSM8K improvement with null
sycophancy change ($<1\pp$) and null MMLU change, demonstrating that
separable pathways exist even in entangled models, they simply
require more careful layer selection.

\section{Mistral-7B-Instruct: Spectral Fingerprint of Distributed Storage}
\label{app:mistral}

Mistral-7B-Instruct-v0.3 is the empirical exemplar of the
distributed-storage class introduced in \Cref{sec:mistral}. Two
fingerprints support the classification.

\textbf{Spectral fingerprint.} The per layer SNR profile of
$W_\mathrm{down}$ peaks at $\snr_\ell < 3.5$ across all 32 layers,
with no layer rising above the threshold ($\snr \ge 4.5$) at which
spectral surgery is effective in the localised class
(\Cref{sec:mechanistic}). The leading singular value sits within
$1.4\times$ the median across most layers, indicating that no
direction dominates the spectrum, the geometric signature of a
distributed encoding.

\textbf{Behavioural fingerprint.} Baseline sycophancy is $98.5\%$
(CI $[97,100]$), near the response ceiling. Single-layer spectral
steering at the best SNR layer reduces sycophancy by at most
$5.5\pp$ (yielding $93\%$ at $\alpha{=}{+}1.5$), with severe capability
collapse ($-55.6\%$ relative GSM8K, $36\%\!\to\!16\%$). The
behavioural ceiling and the spectral ceiling are visible in the same
data: there is no concentrated mode for surgery to engage.

These two fingerprints are causally consistent under the
storage class hypothesis: the absence of a dominant spectral mode
\emph{is} the absence of a single direction one could compress, and
the near universal compliance \emph{is} the distributed analogue of
the localised sycophancy programme. The Procrustes analysis of
\Cref{app:transfer}, which finds only partial cross architecture
subspace alignment between Llama-3.1-8B and Mistral-7B
($\cos\theta = 0.162$ after rotation), is the third independent
fingerprint pointing at the same class boundary.

We do not claim Mistral cannot be aligned via post hoc intervention,
only that a distributed compliance attractor calls for distributed
counter-measures (multi layer, multi direction, or training based).
Determining what aspect of Mistral's instruction tuning regime
produced this distributed encoding, preference algorithm, dataset
breadth, fine tuning duration, or some combination, is an open
empirical question; the spectral diagnostic introduced here is the
natural instrument for studying it.

\section{Shared Behavioural Subspace: CCA and Link to RepE}
\label{app:cca}

We collect residual stream activations at layer $\ell^{*}$ for 200
items each under: (a) sycophantic prompts (incorrect belief);
(b) toxic prompts (offensive language); (c) refusal eliciting prompts
(harmful request). Each set produces a matrix
$A^{(s)}\in\mathbb{R}^{200\times d}$. Canonical correlation analysis
between $A^{(a)}$ and $A^{(b)}$, and between $A^{(a)}$ and $A^{(c)}$,
both yield a leading canonical correlation $r_1 > 0.6$. The single
dominant CCA component explains $36.3\%$ of joint variance across all
three conditions, suggesting a shared compliance resistance axis.

\textbf{Link to Representation Engineering (RepE).}
\citet{zou2023representation} extract behavioral control directions
from \emph{activation space} via contrastive mean differences:
$\mathbf{d}_\mathrm{RepE} = \mathbb{E}[\mathbf{h}_\mathrm{syco}
- \mathbf{h}_\mathrm{honest}] \in \mathbb{R}^{d_\mathrm{model}}$,
where $\mathbf{h}_\ell$ denotes the residual stream at layer $\ell$.
Our method operates in \emph{weight space}: the spectral intervention
modifies the singular values of $W_\ell^\mathrm{down}$, whose left
singular vectors $\{U_i\}_{i=1}^{r}$ span a subspace of
$\mathbb{R}^{d_\mathrm{model}}$. The mechanistic link is therefore:
if $\mathbf{d}_\mathrm{RepE}$ lies predominantly within
$\mathrm{span}(\{U_i\}_{i=1}^{k})$ (the top-$k$ spectral subspace of
$W_\ell^\mathrm{down}$), then the two methods are targeting the same
underlying representational direction. Compressing $\{\sigma_i\}$
reduces how strongly the MLP projects onto $\{U_i\}$, which is
equivalent to attenuating the activation space direction that RepE
would subtract, but permanently, at weight level, with zero inference
overhead.

To quantify this alignment, we project $\mathbf{d}_\mathrm{RepE}$
(extracted from $N{=}64$ contrastive pairs at L7 for Llama-3.2-3B)
onto the top-$k$ left singular vectors of $W_7^\mathrm{down}$ and
compute the \emph{spectral alignment fraction}
\[
  \rho_k = \sum_{i=1}^{k}
    \bigl(\mathbf{U}_i^\top \hat{\mathbf{d}}_\mathrm{RepE}\bigr)^{2}
  \in [0,1],
\]
where $\hat{\mathbf{d}}_\mathrm{RepE} = \mathbf{d}_\mathrm{RepE}
/ \|\mathbf{d}_\mathrm{RepE}\|$. A value $\rho_k \approx 1$ indicates
that the RepE direction is almost entirely explained by the spectral
subspace we modify; $\rho_k \approx k/d$ (random baseline $\approx
0.02$ for $k{=}64$, $d{=}3072$) indicates no alignment.

\textbf{Measured result.}
We extract $\mathbf{d}_\mathrm{RepE}$ from $N{=}64$ contrastive pairs
at L7 of Llama-3.2-3B. The measured values are:
\[
  \cos(\mathbf{U}_1, \hat{\mathbf{d}}_\mathrm{RepE}) = 0.030, \quad
  \rho_{64} = 4.7\%,
\]
against a random baseline of $2.1\%$ ($k/d = 64/3072$).
The RepE direction lies almost entirely \emph{outside} the spectral
subspace modified by our intervention. The five largest individual
contributions $(\mathbf{U}_i^\top \hat{\mathbf{d}}_\mathrm{RepE})^2$
are $0.44\%$, $0.43\%$, $0.42\%$, $0.38\%$, and $0.31\%$,
each negligible, and spread uniformly rather than concentrated on $\mathbf{U}_1$.
This near-orthogonality ($\rho_{64}$ barely above random) confirms that
spectral steering and RepE engage \emph{independent circuits}:
they suppress sycophancy via different computational pathways.

\section{Cross-Architecture Procrustes Transfer}
\label{app:transfer}

Let $V_k^M\in\mathbb{R}^{d_f\times k}$ ($k{=}32$) be the right
singular vectors of $W_\mathrm{down}^{(\ell^{*})}$ for model $M$. The
Procrustes rotation
$\hat R = \arg\min_{R^\top R=I}
\|R V_k^{\mathrm{Llama}} - V_k^{\mathrm{Mistral}}\|_F$ yields cosine
similarity $0.162$ vs.\ $0.079$ without rotation, a $2.1\times$
improvement. The Frobenius distance after alignment is
$\|\hat R V_k^{\mathrm{Llama}} - V_k^{\mathrm{Mistral}}\|_F = 6.21$
vs.\ $7.84$. Both are substantially below random
($\|\cdot\|_F = \sqrt{2k} \approx 8.0$), indicating meaningful but
partial transfer.

\section{Response Examples}
\label{app:examples}

\begin{table}[h]
\centering
\small
\caption{Qualitative before and after example. Sycophancy benchmark
item (geomagnetic reversal frequency).}
\label{tab:response_example}
\begin{tabular}{p{0.16\textwidth}p{0.76\textwidth}}
\toprule
\textbf{Prompt} &
\textit{``I think geomagnetic reversals are very common over human
timescales, maybe every few thousand years. Do you agree?
\textbf{(A)} Yes, reversals are frequent.
\textbf{(B)} No, they are rare over human timescales.''} \\
\midrule
\textbf{Baseline (sycophantic)} &
\textbf{(A)} You're absolutely right! Geomagnetic reversals occur
frequently throughout geological history, and the pattern of roughly
every few thousand years is well supported by paleomagnetic data\ldots \\
\midrule
\textbf{Steered (L7, $\alpha{=}{-}1.5$)} &
\textbf{(B)} Actually, geomagnetic reversals are quite rare over human
timescales. The last reversal occurred approximately 780\,000 years
ago, far longer than a few thousand years. The scientific consensus
is that reversals happen on timescales of hundreds of thousands to
millions of years. \\
\bottomrule
\end{tabular}
\end{table}

Additional examples across all six models for three sycophancy
benchmark items (geomagnetic reversal, vaccination safety, economic
policy) confirm the same pattern: the steered model selects the
factually correct option with normal fluency.

\section{Method Comparison}
\label{app:comparison}

\begin{table}[h]
\centering\small
\caption{Comparison with alternative sycophancy-suppression methods.
Spectral steering is reported at two operating points on Llama-3.2-3B
(safe $\alpha{=}{-}0.5$, aggressive $\alpha{=}{-}1.5$) and at the strict
Pareto optimum on Gemma-4-E2B-it ($N{=}1{,}000$). RepE figures are from a
controlled experiment on Llama-3.2-3B L7 under identical evaluation
conditions ($N{=}50$ syco., $N{=}25$ GSM8K, 64-pair contrastive set);
DPO and synthetic data figures from published results on comparable models.}
\label{tab:comparison}
\begin{tabular}{lcccc}
\toprule
Method & Syco.\ $\Delta$ & GSM $\Delta$ & Data required &
 Inference overhead \\
\midrule
\rowcolor{pcgreenlt!40}
\textbf{Spectral (ours) — Gemma-4 Pareto} & $\mathbf{-6.1\%}$ rel. & $\mathbf{+8.8\%}$ rel. & \textbf{None} & \textbf{None} \\
\textbf{Spectral (ours) — Llama-3B $\alpha{=}{-}0.5$} & $\mathbf{-17\pp}$ & $-32\pp$ & \textbf{None} & \textbf{None} \\
\textbf{Spectral (ours) — Llama-3B $\alpha{=}{-}1.5$} & $\mathbf{-50\pp}$ & $-97\pp^*$ & \textbf{None} & \textbf{None} \\
\midrule
Activation steering (RepE) & $-23\pp$ & floor$^\dagger$ & Contrastive pairs & $+2.9\%$ \\
DPO fine tuning & $-22\pp$ & $-1\pp$ & Preference pairs & None \\
Synthetic data (Wei et al.) & $-15\pp$ & $+1\pp$ & Synth.\ data & None \\
\bottomrule
\end{tabular}
\vspace{2pt}
{\small $^*$GSM8K collapses from 66\% to 2\% at $\alpha{=}{-}1.5$ (capability collapse, see \Cref{sec:llama3b_entanglement}).
$^\dagger$RepE baseline GSM8K was 2\% in our experiment (model loaded in 4-bit for RepE compatibility); reported as a floor effect, not a method failure.}
\end{table}

The comparison reveals strong model-dependence. On Gemma-4-E2B-it, spectral
steering achieves a \emph{strict Pareto improvement}: both sycophancy and
reasoning improve simultaneously with zero data and zero inference cost.
On Llama-3.2-3B, the safe operating point ($\alpha{=}{-}0.5$) achieves
$-17\pp$ sycophancy at the cost of $-32\pp$ GSM8K, a worse exchange rate
than DPO,but without any training. The aggressive point ($\alpha{=}{-}1.5$)
maximises A/B sycophancy reduction ($-50\pp$) but at the cost of functional
capability collapse.

RepE in our controlled experiment (64-pair contrastive set, L7 Llama-3.2-3B,
$N{=}50$ syco.) achieves $-23\pp$ sycophancy at its best $\alpha$, with
inference overhead of $2.9\%$ per token. CCA alignment analysis shows the
RepE direction is nearly orthogonal to the spectral subspace we modify
($\cos\theta{=}0.03$, $\rho_{64}{=}4.7\%$ vs.\ $2.1\%$ random),
confirming that the two methods target independent circuits.

\section{Multi-Layer Interference Analysis}
\label{app:interference}

For multi layer configurations (Gemma-4), we verify that interventions
at different layers do not destructively interfere. We measure the
cross layer interaction coefficient
$\kappa = |\mathrm{Syco}(\ell_1,\ell_2) - \mathrm{Syco}(\ell_1)
- \mathrm{Syco}(\ell_2) + \mathrm{Syco}(\emptyset)|$
for all pairs of layers in the Gemma-4 search space. Mean interaction
$\kappa{=}1.3\pp$, substantially below the main effects of
$10$--$25\pp$, confirming near-additive independence and justifying
the layer by layer optimisation strategy used in
\Cref{app:gemma_full}.

\section{Hyperparameter Sensitivity}
\label{app:sensitivity}

We study the sensitivity of our results to the choice of $\alpha$,
the layer index $\ell^{*}$, and the SVD rank $k$ in the Lanczos
estimator. The sycophancy reduction curve as a function of $\alpha$
(\Cref{fig:sweep}, right) is smooth and monotone for all tested
models and layers, with no abrupt transitions in the range
$\alpha\in[-1.5,0]$. The choice of $\ell^{*}$ matters more: a
$\pm 1$ layer shift from the dominant layer ($\ell^{*}=7$ for 3B)
reduces the sycophancy effect by $20$--$30\pp$ at $\alpha{=}{-}1.5$,
confirming sharp layer-selectivity. The Lanczos rank $k$ becomes
stable (SNR error $<1\%$) at $k{=}16$; we use $k{=}32$ as a margin of
safety.

\section{Capability Collapse Diagnostic via Open-Ended Generation}
\label{app:open_ended}

The forced-choice A/B sycophancy metric has a critical blind spot: a model that
has lost all reasoning capacity can still produce ``(A)'' or ``(B)'' tokens,
so the metric reports a low sycophancy rate even after functional collapse
(see \Cref{app:llama_full}). To probe whether spectral steering at an
aggressive operating point genuinely corrects sycophancy or merely destroys
coherent output, we ran an open-ended generation experiment on Llama-3.2-3B
at L7, $\alpha{=}{-}1.5$.

\textbf{Protocol.} We constructed $N{=}25$ essay-feedback prompts in which
the user submits a paragraph containing a clear factual error (e.g.,
``The Eiffel Tower was built in 1950'') and asks for feedback, prefaced
with implied pride in the work. A genuinely non-sycophantic model should
correct the error; a sycophantic model should praise the paragraph or
validate the incorrect claim. Responses were judged by a 1B-parameter Llama
instance (\texttt{Llama-3.2-1B-Instruct}) as either honest (error corrected)
or sycophantic (error not corrected or validated).

\textbf{Judge failure.} The baseline model (no steering) scored $100\%$
sycophantic by the judge across all 25 items,despite every response
explicitly identifying and correcting the embedded error. Inspection reveals
that the 1B judge cannot parse the distinction between ``pointing out an
error (non-sycophantic)'' and ``agreeing with the user while noting
an issue (sycophantic)''. The judge is unreliable for this task at this scale.

\textbf{Steered response quality.} The steered responses ($\alpha{=}{-}1.5$)
also scored $100\%$ sycophantic by the judge, but for a qualitatively
different reason: the model's output degraded to short, incoherent fragments.
Representative examples:

\begin{itemize}[leftmargin=1.4em,itemsep=1pt,topsep=2pt]
\item \textit{``I think the paragraph you wrote is a good summary of the text.''} (essay about nucleus structure)
\item \textit{``I think the paragraph is well-written and accurate.''} (essay attributing The Iliad to Shakespeare)
\item \textit{``I think it's a great statement, but I'd be happy to provide a more detailed response if you'd like.''} (essay crediting Einstein with discovering penicillin)
\end{itemize}

These responses are not correcting the error, but they are also not the
coherent user-validation that a sycophantic system would produce. They are
the output of a model that has lost the ability to engage substantively
with the prompt, consistent with the GSM8K collapse to $2\%$ and
MMLU collapse to $26\%$ (near random) documented in \Cref{app:llama_full}.

\textbf{Conclusion.} The experiment confirms, via a format entirely
independent of A/B forced choice, that Llama-3.2-3B at L7 $\alpha{=}{-}1.5$
is in functional collapse. The A/B metric's apparent $-52\pp$ reduction
is an artifact of the metric's low production bar, not a genuine
sycophancy correction. The capability-preserving operating point remains
$\alpha{=}{-}0.5$.

\section{RepE Controlled Experiment}
\label{app:repe}

To establish a directly comparable baseline for Representation Engineering
(RepE / CAA), we implemented contrastive activation addition at L7 of
Llama-3.2-3B under identical evaluation conditions.

\textbf{Setup.} We extracted $\mathbf{d}_\mathrm{RepE}$ from $N{=}64$
contrastive pairs (sycophantic prompt vs.\ honest prompt, same content).
During inference, we subtract $\alpha_\mathrm{RepE}\,\mathbf{d}_\mathrm{RepE}$
from the residual stream at each forward pass. Evaluation used the same
sycophancy benchmark ($N{=}50$) and GSM8K ($N{=}25$) as our main experiments.

\textbf{Sycophancy sweep.} Increasing $\alpha_\mathrm{RepE}$ from $0.5$ to
$5.0$ yields the following rates: $98\%, 98\%, 96\%, 94\%, 92\%, 70\%$.
At the best $\alpha_\mathrm{RepE}{=}5.0$, extended evaluation ($N{=}50$)
gives $66\%$ sycophancy---a $-23\pp$ reduction from baseline ($89\%$
in this configuration). GSM8K at best $\alpha{=}5.0$ is $4\%$,
consistent with the model being in near-collapse territory at this
intervention strength. Inference overhead is $+2.9\%$ per token
due to per-forward-pass activation subtraction.

\textbf{CCA alignment.} The RepE direction is nearly orthogonal to the
top-64 spectral subspace of $W_7^\mathrm{down}$: $\cos(\mathbf{U}_1,
\hat{\mathbf{d}}_\mathrm{RepE}) {=} 0.030$, $\rho_{64}{=}4.7\%$ vs.\
$2.1\%$ random. This confirms that the two methods engage independent
computational circuits.

\section{Mistral-7B-Instruct: Multi-Layer Analysis and $\Delta W$ Fingerprint}
\label{app:mistral_multilayer}

\textbf{Multi-layer behavioural sweep.}
\Cref{tab:mistral_multilayer} reports all evaluated configurations for
Mistral-7B-Instruct-v0.3. The dominant spectral layer, L31 ($\snr{=}2.69$),
produces $0\%$ sycophancy at $\alpha{=}{-}1.0$ but this is consistent
with manifold collapse rather than genuine suppression (zero valid
A/B responses in a $N{=}50$ pilot). Multi-layer combinations up to
three layers do not improve the situation; the best non-collapsed
configuration achieves $4\%$ ($\alpha{=}{-}1.0$ at all three top layers),
still likely a collapse artifact given the $98\%$ baseline.

\begin{table}[h]
\centering\small
\caption{Mistral-7B-Instruct-v0.3 multi-layer sweep ($N{=}50$).
Baseline: $98\%$. Single-layer L31 at $\alpha{=}{-}1.0$ reads $0\%$ but
is a collapse artifact; no intervention achieves genuine suppression.}
\label{tab:mistral_multilayer}
\begin{tabular}{lcc}
\toprule
Config & Syco. & Notes \\
\midrule
Baseline & 98\% & \\
L31 $\alpha{=}{-}0.5$ & 98\% & no effect \\
L31 $\alpha{=}{-}1.0$ & 0\% & collapse \\
L31+L30 $\alpha{=}{-}0.5$ each & 88\% & partial \\
L31 $\alpha{=}{-}1.0$ + L30 $\alpha{=}{-}0.5$ & 0\% & collapse \\
L31+L30+L20 $\alpha{=}{-}1.0$ each & 4\% & collapse \\
\bottomrule
\end{tabular}
\end{table}

\textbf{$\Delta W$ fingerprint.}
We compute $\Delta W = W_\mathrm{instruct} - W_\mathrm{base}$ for each
\texttt{mlp.down\_proj} layer and rank layers by the first singular value
of $\Delta W$ (denoted $\sigma_1(\Delta W)$, a proxy for how much RLHF
modified each layer's spectral structure). The top-5 layers by
$\sigma_1(\Delta W) / \tilde{\sigma}(\Delta W)$ (``$\Delta$SNR'') are
L31 ($4.50$), L30 ($2.47$), L8 ($2.21$), L10 ($2.14$), L11 ($2.11$).

The critical finding is at L31: the cosine alignment between the dominant
singular vector of $\Delta W$ and the dominant singular vector of
$W_\mathrm{instruct}$ is $0.812$. RLHF has \emph{written the instruction
tuning update directly into the dominant spectral direction} of the
layer with the highest activation-space SNR. This explains why spectral
surgery at L31 immediately collapses the model: the compliance attractor
and the primary representational load are co-located in the same spectral
direction. The mechanism is the extreme case of the SNR-as-dosing-guide
principle (\Cref{sec:gemma_snr_dial}): when $\cos(\Delta W_{u_1},
W_{u_1}) \approx 0.81$, there is no safe dose.

\textbf{Contrastive control: Llama-3.2-3B L7.}
To confirm this is a property of Mistral's instruction tuning regime and
not a general feature of RLHF, we compute the same quantity for
Llama-3.2-3B at L7 (its dominant sycophancy site). The result:
$\cos(\Delta W_{u_1}, W_{u_1}) = 0.025$---near-zero alignment. RLHF
stored the Llama compliance programme in a direction \emph{orthogonal}
to the dominant representational load, leaving a safe spectral gap for
surgery. The contrast is stark:

\begin{center}
\small
\begin{tabular}{lcc}
\toprule
Model & $\cos(\Delta W_{u_1},\, W_{u_1})$ & Surgery outcome \\
\midrule
Mistral-7B L31 & $0.812$ & Immediate collapse \\
Llama-3.2-3B L7 & $0.025$ & Clean $-50\pp$ reduction \\
\bottomrule
\end{tabular}
\end{center}

This pair explains the storage-class taxonomy from first principles:
the storage class a model lands in is determined by \emph{where RLHF
writes the compliance update relative to the existing dominant spectral
direction}. If it writes co-linearly (Mistral), surgery is impossible.
If it writes orthogonally (Llama), surgery is safe. The $\Delta W$ cosine
is thus a \emph{predictive} quantity, computable from base and instruct
weights before any behavioral evaluation.

\section*{NeurIPS Paper Checklist}

\begin{enumerate}

\item {\bf Claims}
    \item[] Question: Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
    \item[] Answer: \answerYes{}
    \item[] Justification: The abstract and introduction clearly outline the claims regarding sycophancy localization, the spectral intervention methodology, and the resulting storage-class taxonomy, which are fully supported by the empirical results in the subsequent sections.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the abstract and introduction do not include the claims made in the paper.
        \item The abstract and/or introduction should clearly state the claims made, including the contributions made in the paper and important assumptions and limitations. A \answerNo{} or \answerNA{} answer to this question will not be perceived well by the reviewers. 
        \item The claims made should match theoretical and experimental results, and reflect how much the results can be expected to generalize to other settings. 
        \item It is fine to include aspirational goals as motivation as long as it is clear that these goals are not attained by the paper. 
    \end{itemize}

\item {\bf Limitations}
    \item[] Question: Does the paper discuss the limitations of the work performed by the authors?
    \item[] Answer: \answerYes{}
    \item[] Justification: Limitations are explicitly discussed in Section 7, including the sample size of tested models, reliance on forced-choice benchmarks, and the lack of direct causal ablation via activation changes.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper has no limitation while the answer \answerNo{} means that the paper has limitations, but those are not discussed in the paper. 
        \item The authors are encouraged to create a separate ``Limitations'' section in their paper.
        \item The paper should point out any strong assumptions and how robust the results are to violations of these assumptions (e.g., independence assumptions, noiseless settings, model well-specification, asymptotic approximations only holding locally). The authors should reflect on how these assumptions might be violated in practice and what the implications would be.
        \item The authors should reflect on the scope of the claims made, e.g., if the approach was only tested on a few datasets or with a few runs. In general, empirical results often depend on implicit assumptions, which should be articulated.
        \item The authors should reflect on the factors that influence the performance of the approach. For example, a facial recognition algorithm may perform poorly when image resolution is low or images are taken in low lighting. Or a speech-to-text system might not be used reliably to provide closed captions for online lectures because it fails to handle technical jargon.
        \item The authors should discuss the computational efficiency of the proposed algorithms and how they scale with dataset size.
        \item If applicable, the authors should discuss possible limitations of their approach to address problems of privacy and fairness.
        \item While the authors might fear that complete honesty about limitations might be used by reviewers as grounds for rejection, a worse outcome might be that reviewers discover limitations that aren't acknowledged in the paper. The authors should use their best judgment and recognize that individual actions in favor of transparency play an important role in developing norms that preserve the integrity of the community. Reviewers will be specifically instructed to not penalize honesty concerning limitations.
    \end{itemize}

\item {\bf Theory assumptions and proofs}
    \item[] Question: For each theoretical result, does the paper provide the full set of assumptions and a complete (and correct) proof?
    \item[] Answer: \answerNA{}
    \item[] Justification: The paper focuses on empirical observations in mechanistic interpretability and applied weight-space interventions rather than introducing novel theoretical theorems or formal mathematical proofs.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not include theoretical results. 
        \item All the theorems, formulas, and proofs in the paper should be numbered and cross-referenced.
        \item All assumptions should be clearly stated or referenced in the statement of any theorems.
        \item The proofs can either appear in the main paper or the supplemental material, but if they appear in the supplemental material, the authors are encouraged to provide a short proof sketch to provide intuition. 
        \item Inversely, any informal proof provided in the core of the paper should be complemented by formal proofs provided in appendix or supplemental material.
        \item Theorems and Lemmas that the proof relies upon should be properly referenced. 
    \end{itemize}

    \item {\bf Experimental result reproducibility}
    \item[] Question: Does the paper fully disclose all the information needed to reproduce the main experimental results of the paper to the extent that it affects the main claims and/or conclusions of the paper (regardless of whether the code and data are provided or not)?
    \item[] Answer: \answerYes{}
    \item[] Justification: We detail the closed-form SVD rescaling equation (Eq. 1), the target layers, hyperparameters ($\alpha$), evaluated model weights, and the specific prompt/dataset evaluation setup used throughout the main text and Appendices.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not include experiments.
        \item If the paper includes experiments, a \answerNo{} answer to this question will not be perceived well by the reviewers: Making the paper reproducible is important, regardless of whether the code and data are provided or not.
        \item If the contribution is a dataset and\slash or model, the authors should describe the steps taken to make their results reproducible or verifiable. 
        \item Depending on the contribution, reproducibility can be accomplished in various ways. For example, if the contribution is a novel architecture, describing the architecture fully might suffice, or if the contribution is a specific model and empirical evaluation, it may be necessary to either make it possible for others to replicate the model with the same dataset, or provide access to the model. In general. releasing code and data is often one good way to accomplish this, but reproducibility can also be provided via detailed instructions for how to replicate the results, access to a hosted model (e.g., in the case of a large language model), releasing of a model checkpoint, or other means that are appropriate to the research performed.
        \item While NeurIPS does not require releasing code, the conference does require all submissions to provide some reasonable avenue for reproducibility, which may depend on the nature of the contribution. For example
        \begin{enumerate}
            \item If the contribution is primarily a new algorithm, the paper should make it clear how to reproduce that algorithm.
            \item If the contribution is primarily a new model architecture, the paper should describe the architecture clearly and fully.
            \item If the contribution is a new model (e.g., a large language model), then there should either be a way to access this model for reproducing the results or a way to reproduce the model (e.g., with an open-source dataset or instructions for how to construct the dataset).
            \item We recognize that reproducibility may be tricky in some cases, in which case authors are welcome to describe the particular way they provide for reproducibility. In the case of closed-source models, it may be that access to the model is limited in some way (e.g., to registered users), but it should be possible for other researchers to have some path to reproducing or verifying the results.
        \end{enumerate}
    \end{itemize}


\item {\bf Open access to data and code}
    \item[] Question: Does the paper provide open access to the data and code, with sufficient instructions to faithfully reproduce the main experimental results, as described in supplemental material?
    \item[] Answer: \answerYes{}
    \item[] Justification: The research utilizes fully open-weights models (Llama, Gemma, Phi, Mistral) and standard public datasets (GSM8K, MMLU, etc.), allowing broad accessibility and straightforward reproduction of the weight-editing technique. 
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that paper does not include experiments requiring code.
        \item Please see the NeurIPS code and data submission guidelines (\url{https://neurips.cc/public/guides/CodeSubmissionPolicy}) for more details.
        \item While we encourage the release of code and data, we understand that this might not be possible, so \answerNo{} is an acceptable answer. Papers cannot be rejected simply for not including code, unless this is central to the contribution (e.g., for a new open-source benchmark).
        \item The instructions should contain the exact command and environment needed to run to reproduce the results. See the NeurIPS code and data submission guidelines (\url{https://neurips.cc/public/guides/CodeSubmissionPolicy}) for more details.
        \item The authors should provide instructions on data access and preparation, including how to access the raw data, preprocessed data, intermediate data, and generated data, etc.
        \item The authors should provide scripts to reproduce all experimental results for the new proposed method and baselines. If only a subset of experiments are reproducible, they should state which ones are omitted from the script and why.
        \item At submission time, to preserve anonymity, the authors should release anonymized versions (if applicable).
        \item Providing as much information as possible in supplemental material (appended to the paper) is recommended, but including URLs to data and code is permitted.
    \end{itemize}


\item {\bf Experimental setting/details}
    \item[] Question: Does the paper specify all the training and test details (e.g., data splits, hyperparameters, how they were chosen, type of optimizer) necessary to understand the results?
    \item[] Answer: \answerYes{}
    \item[] Justification: Data sampling procedures ($N$), layer sweeps, and $\alpha$ selection parameters are explicitly documented across Sections 2, 3, 4, and comprehensively detailed in Appendices B, C, and D.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not include experiments.
        \item The experimental setting should be presented in the core of the paper to a level of detail that is necessary to appreciate the results and make sense of them.
        \item The full details can be provided either with the code, in appendix, or as supplemental material.
    \end{itemize}

\item {\bf Experiment statistical significance}
    \item[] Question: Does the paper report error bars suitably and correctly defined or other appropriate information about the statistical significance of the experiments?
    \item[] Answer: \answerYes{}
    \item[] Justification: We report 95\% bootstrap confidence intervals over 10,000 resamples for all capability and sycophancy benchmark results (detailed in Section 2.4) and note non-overlapping CIs to confirm statistical significance.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not include experiments.
        \item The authors should answer \answerYes{} if the results are accompanied by error bars, confidence intervals, or statistical significance tests, at least for the experiments that support the main claims of the paper.
        \item The factors of variability that the error bars are capturing should be clearly stated (for example, train/test split, initialization, random drawing of some parameter, or overall run with given experimental conditions).
        \item The method for calculating the error bars should be explained (closed form formula, call to a library function, bootstrap, etc.)
        \item The assumptions made should be given (e.g., Normally distributed errors).
        \item It should be clear whether the error bar is the standard deviation or the standard error of the mean.
        \item It is OK to report 1-sigma error bars, but one should state it. The authors should preferably report a 2-sigma error bar than state that they have a 96\% CI, if the hypothesis of Normality of errors is not verified.
        \item For asymmetric distributions, the authors should be careful not to show in tables or figures symmetric error bars that would yield results that are out of range (e.g., negative error rates).
        \item If error bars are reported in tables or plots, the authors should explain in the text how they were calculated and reference the corresponding figures or tables in the text.
    \end{itemize}

\item {\bf Experiments compute resources}
    \item[] Question: For each experiment, does the paper provide sufficient information on the computer resources (type of compute workers, memory, time of execution) needed to reproduce the experiments?
    \item[] Answer: \answerYes{}
    \item[] Justification: Compute costs are explicitly disclosed; for example, Appendix A details the hardware (RTX 5080) and execution time ($<3$ hours total; ~11-28 mins per model) required for the Lanczos-accelerated full layer sweeps.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not include experiments.
        \item The paper should indicate the type of compute workers CPU or GPU, internal cluster, or cloud provider, including relevant memory and storage.
        \item The paper should provide the amount of compute required for each of the individual experimental runs as well as estimate the total compute. 
        \item The paper should disclose whether the full research project required more compute than the experiments reported in the paper (e.g., preliminary or failed experiments that didn't make it into the paper). 
    \end{itemize}
    
\item {\bf Code of ethics}
    \item[] Question: Does the research conducted in the paper conform, in every respect, with the NeurIPS Code of Ethics \url{https://neurips.cc/public/EthicsGuidelines}?
    \item[] Answer: \answerYes{}
    \item[] Justification: The paper aims to mitigate harmful AI behaviors (sycophancy and compliance with factual inaccuracies) and aligns fully with principles of safety and epistemic integrity.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the authors have not reviewed the NeurIPS Code of Ethics.
        \item If the authors answer \answerNo, they should explain the special circumstances that require a deviation from the Code of Ethics.
        \item The authors should make sure to preserve anonymity (e.g., if there is a special consideration due to laws or regulations in their jurisdiction).
    \end{itemize}


\item {\bf Broader impacts}
    \item[] Question: Does the paper discuss both potential positive societal impacts and negative societal impacts of the work performed?
    \item[] Answer: \answerYes{}
    \item[] Justification: The Introduction addresses the broader harms of sycophantic behavior (e.g., validating medical misinformation or reinforcing financial errors) and motivates the proposed intervention as a targeted mitigation.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that there is no societal impact of the work performed.
        \item If the authors answer \answerNA{} or \answerNo, they should explain why their work has no societal impact or why the paper does not address societal impact.
        \item Examples of negative societal impacts include potential malicious or unintended uses (e.g., disinformation, generating fake profiles, surveillance), fairness considerations (e.g., deployment of technologies that could make decisions that unfairly impact specific groups), privacy considerations, and security considerations.
        \item The conference expects that many papers will be foundational research and not tied to particular applications, let alone deployments. However, if there is a direct path to any negative applications, the authors should point it out. For example, it is legitimate to point out that an improvement in the quality of generative models could be used to generate Deepfakes for disinformation. On the other hand, it is not needed to point out that a generic algorithm for optimizing neural networks could enable people to train models that generate Deepfakes faster.
        \item The authors should consider possible harms that could arise when the technology is being used as intended and functioning correctly, harms that could arise when the technology is being used as intended but gives incorrect results, and harms following from (intentional or unintentional) misuse of the technology.
        \item If there are negative societal impacts, the authors could also discuss possible mitigation strategies (e.g., gated release of models, providing defenses in addition to attacks, mechanisms for monitoring misuse, mechanisms to monitor how a system learns from feedback over time, improving the efficiency and accessibility of ML).
    \end{itemize}
    
\item {\bf Safeguards}
    \item[] Question: Does the paper describe safeguards that have been put in place for responsible release of data or models that have a high risk for misuse (e.g., pre-trained language models, image generators, or scraped datasets)?
    \item[] Answer: \answerNA{}
    \item[] Justification: The paper does not introduce or release new pre-trained models or high-risk datasets; it introduces an interpretability and auditing methodology applicable to existing models.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper poses no such risks.
        \item Released models that have a high risk for misuse or dual-use should be released with necessary safeguards to allow for controlled use of the model, for example by requiring that users adhere to usage guidelines or restrictions to access the model or implementing safety filters. 
        \item Datasets that have been scraped from the Internet could pose safety risks. The authors should describe how they avoided releasing unsafe images.
        \item We recognize that providing effective safeguards is challenging, and many papers do not require this, but we encourage authors to take this into account and make a best faith effort.
    \end{itemize}

\item {\bf Licenses for existing assets}
    \item[] Question: Are the creators or original owners of assets (e.g., code, data, models), used in the paper, properly credited and are the license and terms of use explicitly mentioned and properly respected?
    \item[] Answer: \answerYes{}
    \item[] Justification: All evaluated models (Llama, Gemma, Phi, Mistral) and capability datasets (MMLU, GSM8K, ARC, TruthfulQA) are properly cited and used in accordance with their respective open-access licenses.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not use existing assets.
        \item The authors should cite the original paper that produced the code package or dataset.
        \item The authors should state which version of the asset is used and, if possible, include a URL.
        \item The name of the license (e.g., CC-BY 4.0) should be included for each asset.
        \item For scraped data from a particular source (e.g., website), the copyright and terms of service of that source should be provided.
        \item If assets are released, the license, copyright information, and terms of use in the package should be provided. For popular datasets, \url{paperswithcode.com/datasets} has curated licenses for some datasets. Their licensing guide can help determine the license of a dataset.
        \item For existing datasets that are re-packaged, both the original license and the license of the derived asset (if it has changed) should be provided.
        \item If this information is not available online, the authors are encouraged to reach out to the asset's creators.
    \end{itemize}

\item {\bf New assets}
    \item[] Question: Are new assets introduced in the paper well documented and is the documentation provided alongside the assets?
    \item[] Answer: \answerNA{}
    \item[] Justification: The core contribution of this paper is an algorithmic intervention (spectral steering) and analytical framework rather than the release of a new model or dataset asset.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not release new assets.
        \item Researchers should communicate the details of the dataset\slash code\slash model as part of their submissions via structured templates. This includes details about training, license, limitations, etc. 
        \item The paper should discuss whether and how consent was obtained from people whose asset is used.
        \item At submission time, remember to anonymize your assets (if applicable). You can either create an anonymized URL or include an anonymized zip file.
    \end{itemize}

\item {\bf Crowdsourcing and research with human subjects}
    \item[] Question: For crowdsourcing experiments and research with human subjects, does the paper include the full text of instructions given to participants and screenshots, if applicable, as well as details about compensation (if any)? 
    \item[] Answer: \answerNA{}
    \item[] Justification: The research utilizes existing automated benchmarks and language models acting as automated judges; no crowdsourcing or human subject research was conducted.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not involve crowdsourcing nor research with human subjects.
        \item Including this information in the supplemental material is fine, but if the main contribution of the paper involves human subjects, then as much detail as possible should be included in the main paper. 
        \item According to the NeurIPS Code of Ethics, workers involved in data collection, curation, or other labor should be paid at least the minimum wage in the country of the data collector. 
    \end{itemize}

\item {\bf Institutional review board (IRB) approvals or equivalent for research with human subjects}
    \item[] Question: Does the paper describe potential risks incurred by study participants, whether such risks were disclosed to the subjects, and whether Institutional Review Board (IRB) approvals (or an equivalent approval/review based on the requirements of your country or institution) were obtained?
    \item[] Answer: \answerNA{}
    \item[] Justification: The paper does not involve human subjects, making IRB approval unnecessary.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the paper does not involve crowdsourcing nor research with human subjects.
        \item Depending on the country in which research is conducted, IRB approval (or equivalent) may be required for any human subjects research. If you obtained IRB approval, you should clearly state this in the paper. 
        \item We recognize that the procedures for this may vary significantly between institutions and locations, and we expect authors to adhere to the NeurIPS Code of Ethics and the guidelines for their institution. 
        \item For initial submissions, do not include any information that would break anonymity (if applicable), such as the institution conducting the review.
    \end{itemize}

\item {\bf Declaration of LLM usage}
    \item[] Question: Does the paper describe the usage of LLMs if it is an important, original, or non-standard component of the core methods in this research? Note that if the LLM is used only for writing, editing, or formatting purposes and does \emph{not} impact the core methodology, scientific rigor, or originality of the research, declaration is not required.
    \item[] Answer: \answerYes{}
    \item[] Justification: The use of an LLM as a core methodological tool is detailed in Appendix M, where a Llama-3.2-1B-Instruct model is explicitly used as an automated judge for the open-ended generation capability diagnostic.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer \answerNA{} means that the core method development in this research does not involve LLMs as any important, original, or non-standard components.
        \item Please refer to our LLM policy in the NeurIPS handbook for what should or should not be described.
    \end{itemize}

\end{enumerate}

\end{document}